The common TMEM173 HAQ, AQ alleles rescue CD4 T cellpenia, restore T-regs, and prevent SAVI (N153S) inflammatory disease in mice

Version 1

Abstract

Abstract

The significance of STING (encoded by the TMEM173 gene), in tissue inflammation and cancer immunotherapy has been increasingly recognized. Intriguingly, common human STING alleles R71H-G230A-R293Q (HAQ) and G230A-R293Q (AQ) are carried by ∼60% of East Asians and ∼40% of Africans, respectively. Here, we examine the modulatory effects of HAQ, AQ alleles on STING-associated vasculopathy with onset in infancy (SAVI), an autosomal dominant, fatal inflammatory disease caused by gain-of-function human STING mutations. CD4 T cellpenia is evident in SAVI patients and mouse models. Using STING knock-in mice expressing common human STING alleles HAQ, AQ, and Q293, we found that HAQ, AQ, and Q293 splenocytes resist STING-mediated cell death ex vivo, establishing a critical role of STING residue 293 in cell death. The HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice did not have CD4 T cellpenia. The HAQ/SAVI(N153S), AQ/SAVI(N153S) mice have more (∼10-fold, ∼20-fold, respectively) T-regs than WT/SAVI(N153S) mice. Remarkably, while they have comparable TBK1, IRF3, and NFκB activation as the WT/SAVI, the AQ/SAVI mice have no tissue inflammation, regular body weight, and normal lifespan. We propose that STING activation promotes tissue inflammation by depleting T-regs cells in vivo. Billions of modern humans have the dominant HAQ, AQ alleles. STING research and STING-targeting immunotherapy should consider TMEM173 heterogeneity in humans.

Teaser: One copy of the common human HAQ or AQ TMEM173 gene prevents the gain-of-function TMEM173 mutant-caused SAVI disease in mice.

Introduction

STING drives cytosolic DNA-induced type I IFNs production 1. Recent research revealed that STING promotes inflammation in a variety of inflammatory diseases including nonalcoholic fatty liver disease, nonalcoholic steatohepatitis, kidney injury, neurodegenerative diseases, cardiovascular diseases, obesity, diabetes, and aging 2–9. The type I IFNs-independent function of STING has also emerged 9–12. For example, initially described as a type I interferonopathy 13, recent studies in STING-associated vasculopathy with onset in infancy (SAVI) mouse models showed that SAVI is largely independent of type I IFNs 14–17. In a STING N153S mouse model of SAVI, crossing N153S mice to IRF3/IRF7, and IFNAR1 knockout mice, N153S mice still developed spontaneous lung diseases 14. In another example of STING-dependent, type I IFNs-independent inflammatory disease, DNase II-deficient mice developed autoinflammatory arthritis that can be rescued by STING deficiency 18. But Dnase II−/− mice with the STING-S365A mutation still exhibited severe polyarthritis 19,20. The STING-S365A mutant cannot recruit IRF3 and produce IFNβ 10,11,19,20. How STING drives inflammation in vivo, independent of type I IFNs, remains unknown. The activation of NFκB or NLRP3 inflammasome by STING has been proposed 9,21.

Characterized as an innate immune sensor, STING expression is, paradoxically, high in CD4 T cells 13,22. Furthermore, STING activation kills mouse and human CD4 T cells ex vivo 23–25. SAVI patients and mouse model had CD4 T cellpenia 13,25. STING was first discovered as MPYS for its cell growth inhibition and cell death function in mouse B lymphoma cells 26. STING-mediated cell death is cell type dependent. For example, while STING activation kills human endothelial cells, primary and cancerous T cells, it does not kill mouse MEFs, BMDCs, or BMDMs 27–29. Second, STING-mediated cell death is type I IFNs-independent 25,27,29. Multiple cell death pathways, i.e., apoptosis, necroptosis, pyroptosis, ferroptosis, and PANoptosis, are proposed 29–31. Last, the in vivo biological significance of STING-mediated CD4 T cell death is not clear 25,29. WT and STING−/− mice have similar thymic and pLN CD4−CD8−, CD4+CD8+, and single positive CD4 and CD8 T cell populations 23,24. The expression of CD4 and CD8 TCR are comparable 23,24. The naive, memory, or regulatory T cell populations are similar in WT and STING−/− mice that expressed comparable levels of CD69 and CD25 following TCR activation 23,24. Together, the absence of STING in mice does not alter T cell development or TCR signaling 23. In humans, SAVI patients with constitutively activated STING have low CD4 T cell numbers 13, and type I IFNs are dispensable for STING-mediated human CD4 T cell death 25. Notably, SAVI patients (N154S or V155M) had normal counts of CD8 T and B cells 13.

The human TMEM173 gene is highly heterogeneous 32,33. ∼50% of people in the U.S. carry at least one copy of non-WT TMEM173 allele 32. Among them, the R71H-G230A-R293Q (HAQ) is the 2nd most common TMEM173 allele carried by ∼23% of people in the U.S. 32. However, in East Asians, WT/HAQ (34.3%), not WT/WT (22.0%), is the most common TMEM173 genotype 33. Critically, the HAQ allele was positively selected in modern humans outside Africa 34. Anatomically modern humans outside Africa are descendants of a single Out-of-Africa Migration 50,000∼70,000 years ago. ∼1.4% of Africans have the HAQ allele, while ∼63.9% of East Asians are HAQ carriers 34. Haplotype analysis revealed that HAQ was derived from G230A-R293Q (AQ) allele 34. Importantly, the AQ allele was negatively selected outside Africa 34. ∼40.1% of Africans are AQ carriers, while ∼0.4% of East Asians have the AQ allele 34. Importantly, TMEM173 alleles often have a dominant negative effect likely because the protein STING exists as a homodimer 26,35. SAVI is an autosomal dominant inflammatory disease 13. WT/HAQ individuals had reduced Pneumovax23®-induced antibody responses compared to WT/WT individuals (NCT02471014) 36. Notably, AQ responds to CDNs and produce type I IFNs in vivo and in vitro 34,37,38, but the AQ allele was negatively selected in non-Africans 34. In contrast, the HAQ allele, defective in CDNs-type I IFNs responses 32,33,36,39,40, was positively selected in non-Africans 34 indicating that the CDNs-type I IFNs independent function of STING was essential for the survival of early modern humans outside of Africa.

In this study, we discovered, surprisingly, that the HAQ, AQ splenocytes are resistant to STING-mediated cell death. We generated HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice and found that the HAQ, AQ alleles prevent CD4 T cellpenia, increase/restore T-regs and alleviate/stop tissue inflammation in SAVI mice, thus providing evidence for the in vivo significance of type I IFNs-independent, STING-mediated CD4 T cell death.

Results

STING activation kills mouse spleen CD4, CD8 T and CD19 B cells ex vivo

We first used the synthetic non-CDNs STING agonist diABZI 41 to induce lymphocyte death because diABZI induces cell death without the needs for lipid transfection or detergent for cell permeabilization 28,42 and diABZI is in clinical trials (NCT05514717). Splenocytes from C57BL/6N mice were treated with diABZI in culture, and cell death was determined by Annexin V and Propidium Iodide stain. Splenocyte cell death could be detected as early as 5hrs post diABZI treatment (Figure S1A). Dosage responses showed that ∼25ng/ml diABZI could kill 70% of splenocytes (Figure S1B). Similarly, STING agonists DMXAA, and synthetic CDNs RpRpssCDA, killed mouse spleen CD4, CD8 T cells, and CD19 B cells (Figure 1A). Thus, STING activation readily induces mouse lymphocyte death ex vivo.

Splenocytes from HAQ, AQ, and Q293 mice are resistant to STING-mediated cell death ex vivo.

(A). C57BL/6N splenocytes were treated directly (no transfection) with diABZI (100ng/ml), RpRpssCDA (5 μg/ml) or 2′3′-cGAMP (10μg/ml), DMXAA (25μg/ml) for 24hrs in culture. CD4, CD8 T cells and CD19 B cells death were determined by PI staining. (B). Splenocytes from C57BL/6N mice were pre-treated with indicated small molecules, GSK2334470 (1.25µM), GSK8612 (2.5µM), Bx-795 (0.5µM), QVD-OPH (25µM) for 2hrs. diABZI (100ng/ml) was added in culture for another 24hrs. Dead cells were determined by PI staining. (C-D). Flowcytometry of HAQ, AQ, IFNAR1-/- or C57BL/6N splenocytes treated with diABZI (100ng/ml) for 24hrs. Cell death was determined by PI staining. (E-F). Q293 or the WT littermates splenocytes were treated with diABZI (100ng/ml), RpRpssCDA (5 μg/ml) or 2′3′-cGAMP 10μg/ml) for 24h. Cell death was determined by PI staining. (G). WT/HAQ, WT/AQ, or WT/WT littermates splenocytes were treated with DMXAA (10, 25 or 100µg/ml) for 24h. Cell death was determined by PI staining. Data are representative of three independent experiments. Graphs represent the mean with error bars indication s.e.m. p values are determined by one-way ANOVA Tukey’s multiple comparison test (A, E, G) or unpaired student T-test (B, D). * p<0.05. n.s: not significant.

TBK1 activation is required for STING-mediated mouse spleen cell death ex vivo

STING activation can lead to apoptosis, pyroptosis, necroptosis, or ferroptosis 25–31,42,43. We then treated mouse splenocytes with apoptosis, pyroptosis, necroptosis inhibitors, STING inhibitors H-151, C-176, palmitoylation inhibitor 2-bromopalmitate (2-BP), followed by diABZI stimulation. Inhibitors for NLRP3 (MCC950), RIPK1 (Necrostatin-1), RIPK3 (GSK872), Caspase-1 (VX-795), Caspase-3 (Z-DEVD-FMK), Caspase 1,3,8,9 (Q-VD-Oph), ferroptosis (liproxstatin-1) did not affect diABZI-induced splenocyte cell death ex vivo (Figure S1B, S1C). The STING inhibitors H-151, C-176, and 2-BP also could not prevent diABZI-induced cell death (Figure S1C) though they inhibited diABZI-induced IFNβ production (Figure S1D). Instead, the TBK1 inhibitor BX-795 abolished diABZI-induced splenocyte death (Figure S1C).

BX-795 is a multi-kinase inhibitor including 3-phosphoinositide-dependent protein kinase 1 (PDK1) and TBK1 (IC50s = 6 and 11 nM, respectively). However, the treatment of PDK1 inhibitor GSK2334470 (IC50 = 10nM) did not prevent diABZI-induced splenocyte death (Figure 1B). In contrast, GSK8612, a highly potent and selective inhibitor for TBK1, prevented diABZI-induced splenocyte death (Figure 1B). Thus, TBK1 activation is likely critical for STING-mediated lymphocyte cell death ex vivo.

HAQ, AQ, Q293 STING knock-in mouse splenocytes are resistant to STING-mediated cell death ex vivo

HAQ and AQ are common human TMEM173 alleles 32–34. Previously, we reported that HAQ knock-in mice are defective in CDNs-induced immune responses, while CDNs responses in AQ knock-in mice are similar to WT mice 34. We treated splenocytes from HAQ and AQ mice with diABZI ex vivo and found, surprisingly, that both HAQ and AQ splenocytes were resistant to diABZI-induced cell death (Figure 1C, 1D). In comparison, IFNAR1-/- splenocytes were killed by diABZI, confirming that STING-mediated lymphocytes death are type I IFNs-independent (Figure 1C, 1D) 25,27,29.

HAQ and AQ share the common A230 and Q293 residues changes. We thus generated a Q293 TMEM173 knock-in mouse. Notably, the Q293 splenocytes were resistant to STING agonists 2’3’-cGAMP, RpRpssCDA, and diABZI-induced cell death (Figure 1E, 1F). Thus, the residue 293 of STING is critical for its cell death function.

WT/HAQ, WT/AQ mouse splenocytes are partially resistant to STING-mediated cell death ex vivo

WT/HAQ (34.3%), not WT/WT (22.0%), is the most common human STING genotype in East Asians, while WT/AQ (28.2%) is the 2nd most common STING genotype in Africans 33. We generated WT/HAQ, WT/AQ mice and treated the splenocytes with mouse STING agonist DMXAA. WT/HAQ and WT/AQ splenocytes were protected from 25µg/ml DMXAA-induced cell death (Figure 1G). 100µg/ml DMXAA could kill WT/HAQ and WT/AQ splenocytes albeit less than WT/WT cells (Figure 1G). Thus, the HAQ and AQ alleles are dominant and likely impact STING activation even in heterozygosity.

STING activation kills primary human CD4 T cells, but not CD8 T or CD19 B cells

STING agonists-based clinical trials in humans have been disappointing (NCT02675439, NCT03010176, NCT05514717) 44,45. We showed that the human TMEM173 gene might undergo natural selection during the out-of-Africa migration 34 sensitive to evolutionary pressure. Thus, we investigated STING-mediated death in primary human lymphocytes.

Human explant lung cells from the WT(R232)/WT(R232) donors were treated with STING agonists 2’3-c GAMP, RpRpssCDA, diABZI for 24hrs in culture. Lymphocyte cell death was determined by Propidium Iodide staining. Different from mouse lymphocytes, diABZI and RpRpssCDA killed human CD4 T but not CD8 T or CD19 B cells (Figure 2A, 2B). Human CD8 T and CD19 B cells are resistant to 500ng/ml diABZI-induced cell death (Figure S2A).

HAQ, AQ, and Q293 human cells are resistant to STING agonists-induced death.

(A-B). Total human Lung cells from WT/WT individuals were treated with diABZi (100ng/ml) for 24hrs. Cell death in CD4, CD8 T cells and CD19 B cells were determined by PI staining. (C). Total lung cells from a WT/HAQ (2 individuals) and a WT/WT (3 individuals) were treated with diABZi (25, 100ng/ml) for 24h. Cell death in CD4 T cells was determined by PI staining. (D-E). STING-KO THP-1 cells (Invivogen,, cat no. thpd-kostg) were stably reconstituted with human WT (R232), HAQ, AQ, Q293. Cells were treated with diABZI (200ng/ml) in culture for 24 hrs. Dead cells were determined by Annexin V staining. (F). STING-KO THP-1 cells stably reconstituted with human WT (R232), Q293 were treated with indicated dose of diABZI for 24hs in culture. Dead cells were determined by Annexin V staining. (G). STING-KO THP-1 cells stably reconstituted with human WT (R232), Q293 were treated with indicated dose of diABZI for 2hs in culture. STING activation was detected by anti-STING antibody (Proteintech, #19851-1-AP). (H). STING-KO THP-1 cells stably reconstituted with human WT (R232), HAQ, AQ, Q293 were treated with 50ng/ml diABZI in culture for 24hrs. ISG-54 reporter luciferase activity was determined in cell supernatant and normalized to 10ng/ml IFNβ-stimulated ISG-54 luciferase activity. (I). STING-KO THP-1 cells stably reconstituted with human WT (R232), HAQ, AQ were treated with 50ng/ml diABZI in culture for 2hrs. STING and IRF3 activation were determined by anti-STING antibody and anti-p IRF3 antibody (CST, Ser396, clone 4D4G). Densitometry was determined by ImageLab 5. Data are representative of three independent experiments. Graphs represent the mean with error bars indication s.e.m. p values determined by one-way ANOVA Tukey’s multiple comparison test (B, C, F, H, G) or unpaired student T-test (E, I). * p<0.05, n.s: not significant.

WT/HAQ human CD4 T cells are resistant to low-dose of diABZI-induced cell death

WT/HAQ mouse splenocytes are resistant to low dose diABZI-induced cell death. To extend our observation into primary human T cells, we obtained lung explants from WT/WT and WT/HAQ individuals (Figure S2B) and treated them with diABZI in culture. 25ng/ml diABZI killed WT/WT, but not WT/HAQ, human lung CD4 T cells (Figure 2C).

diABZI induces cell death in STING-KO human THP-1 cells reconstituted with WT human STING (R232) but not HAQ, AQ or Q293 human STING allele

To further determine cell death influenced by human TMEM173 alleles HAQ, AQ and Q293, we used the STING-KO THP-1 cell line because STING agonist induces type I IFNs and cell death in STING-KO THP-1 cells expressing WT human STING 31 (Figure S2C, S2D). We, thus, generated stable THP-1 STING-KO lines expressing HAQ, AQ, WT or Q293 TMEM173 allele. Cell death was determined by Annexin-V staining. diABZI killed THP-1 STING-KO lines expressing WT but not HAQ, AQ, or Q293 TMEM173 allele (Figure 2D, 2E). No cell death was induced in the Q293 THP-1 cells stimulated by 20 to 200 ng/ml of diABZI (Figure 2F). diABZI also did not induce STING activation in Q293 THP-1 cells (Figure 2G). Notably, 50ng/ml diABZI induced p-IRF3 activation and type I IFNs in AQ THP-1 cells, but not HAQ THP-1 cells (Figure 2H, 2I) indicating that the STING-cell death and STING-IRF3-Type I IFNs pathways can be uncoupled.

HAQ and AQ alleles rescue the lymphopenia and suppress myeloid cell expansion in SAVI(N153S) mice

The in vivo significance of the STING/MPYS-cell death is unclear. Furthermore, multiple cell death pathways, i.e., apoptosis, necroptosis, pyroptosis, ferroptosis, and PANoptosis, are proposed 29–31. The uncertainty likely results from studies using different cell types (primary cells vs cancer cell lines); species (human vs mouse); STING agonists (cGAMP, which requires cell permeabilization by detergents or lipid transfection, vs diABZi, DMXAA that can directly cross the membrane) 23,24,27,28,46. Critically, which mechanism is relevant in vivo, causing T cellpenia is not known. To clarify the in vivo significance and mechanisms of STING-mediated cell death, we turned to SAVI mice.

SAVI is an autosomal dominant, inflammatory disease caused by one copy of a gain-of-function STING mutant 13. CD4 T cellpenia was found in SAVI patients and SAVI mouse models 13,14. STING activation in SAVI mice is independent of ligands and happens in vivo. We thus generated HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice aiming to establish the in vivo significance and mechanism of STING-cell death.

First, HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice had reduced splenomegaly compared to WT/SAVI(N153S) mice though their spleens were still larger than the littermates WT/HAQ and WT/AQ (Figure 3A, 3B). Next, HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice had similar spleen B cells and CD4 T cells numbers as the WT/HAQ, WT/AQ littermates (Figure 2B, 2C). Their CD8+ T cells were lower than their WT littermates but much higher than the WT/SAVI(N153S) mice (Figure 3D). Third, spleen myeloid cell numbers, i.e., neutrophils, Ly6Chi monocytes and F4/80 macrophages, were all reduced by half compared to WT/SAVI(N153S) mice (Figure 3E∼3H). Notably, the HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice also had restored bone marrow monocyte (Figure S3). Thus, HAQ and AQ alleles prevent lymphopenia and suppress myeloid cells expansion in SAVI(N153S) mice.

HAQ and AQ rescue the lymphopenia and suppress myeloid cell expansion in SAVI (N153S) mice.

(A). The size and weight of spleens from WT/HAQ, HAQ/SAVI, AQ/SAVI, WT/AQ, WT/SAVI. (B-D). Spleen CD19+ B cells, CD4, CD8 T cells were determined in the indicated mice by Flow. (E-H). Spleen Ly6G+ neutrophils, Ly6Chi monocytes and F4/80+ macrophage was determined in the indicated mice by Flow. Data are representative of three independent experiments. n=3∼5 mice/group. Graphs represent the mean with error bars indication s.e.m. P values are determined by one-way ANOVA Tukey’s multiple comparison test. * p<0.05, n.s: not significant.

The HAQ allele alleviates and the AQ allele prevents SAVI(N153S) disease in mice

SAVI(N153S) disease is characterized by early onset, failure to thrive (low body weight), persistent lung inflammation, decreased lung function, and young death in humans and mouse models 13,17,46,47. The HAQ/SAVI(N153S) mice weighed more and had an improved lifespan than the WT/SAVI(N153S) mice (Figure 4A, 4B). The lifespan, airway resistance and tissue inflammation (lung, liver) were also improved in HAQ/SAVI(N153S) mice compared to the WT/SAVI(N153S) mice (Figure 4C, 4D, 4J, 4K, S4). However, the pulmonary artery pressure was still elevated in HAQ/SAVI(N153S) mice (Figure 4E). Remarkably, the AQ/SAVI(N153S) mice had similar body weight and lifespan as the WT/AQ mice (Figure 4F, 4G). The airway resistance, pulmonary artery pressure, and tissue inflammation in AQ/SAVI(N153S) were similar to the WT/AQ littermates (Figure 4H, 4I, 4J, 4K, S4). Thus, the AQ allele prevents inflammatory SAVI disease in mice.

HAQ and AQ alleles prevent SAVI(N153S) disease in mice.

(A, B, G). The size and body weight of HAQ/SAVI, AQ/SAVI, WT/SAVI and their littermates WT/HAQ, WT/AQ mice. (D, E, H, I). Airway resistance, and pulmonary artery pressure were determined as described in Materials and Methods. (C, F). HAQ/SAVI, AQ/SAVI, WT/SAVI (10 mice/group), were monitored for survival by Kaplan-Meier. (J, K). Representative hematoxylin and eosin (H&E) staining of lung, liver sections from indicated mice. n=3∼5 mice/group. Data are representative of 3 independent experiments. Graphs represent the mean with error bars indication s.e.m. p values are determined by one-way ANOVA Tukey’s multiple comparison test. * p<0.05, **p<0.01. n.s.: not significant. WBC: white blood cells; H: hepatocytes; K: Kupper cells; PV: portal vein; CV: central vein.

diABZI induces similar STING, TBK1, IRF3, NFκB activation in the AQ/SAVI(N153S) and WT/SAVI(N153S) BMDM

SAVI was characterized as type I interferonopathy 17. However, several studies showed that type I IFN signaling and IRF3 activation were dispensable for SAVI disease 14,15,47,48. AQ allele prevents SAVI disease (Figure 4). However, diABZI-treated AQ/SAVI(N153S) and WT/SAVI(N153S) BMDM had similar TBK1-IRF3 activation and IFNβ production (Figure 5A, 5D). diABZI treatment caused IκBα degradation, and similar TNF production in WT/SAVI and AQ/SAVI BMDM (Figure 5B, 5E). Furthermore, diABZI activation led to STING protein degradation in WT/SAVI and AQ/SAVI BMDM (Figure 5C). Last, using cleavable crosslinker dithiobis (succinimidyl propionate (DSP), we showed that STING in WT/SAVI, AQ/SAVI BMDM forms similar dimer in situ (Figure 5F). Thus, the AQ/SAVI BMDM had similar STING degradation, TBK1, IRF3, NFκB activation, and dimerization as the WT/SAVI BMDM.

AQ/SAVI(N153S) cells had similar TBK1-IRF3, NFκB activation and STING degradation as the WT/SAVI(N153S) cells.

(A, B). BMDM from WT/WT, WT/SAVI, HAQ/SAVI and AQ/SAVI were treated with 100ng/ml diABZi in culture for 2 hrs. Cells were lysed and run on a 4∼20% Mini-PROTEAN TGX Stain-Free Precast Gel. The blot was probed for phospho-Thr172-TBK1 antibody (CST, clone D52C2), phosphor-Ser 396-IRF3 (CST, clone 4D4G), STING (Proteintech, #19851-1-AP) and IκBα (CST, clone 44D4) antibody. (C). BMDM from WT/WT, WT/SAVI, HAQ/SAVI and AQ/SAVI were treated with 100ng/ml diABZi in culture for 24 hrs. Cells were lysed and run on a 4∼20% Mini-PROTEAN TGX Stain-Free Precast Gel. The blot was probed for STING antibody (Proteintech, 19851-1-AP). (D, E). IFNβ and TNF were determined by ELISA in the cell supernatant from C. (F). BMDM from WT/WT, WT/SAVI, HAQ/SAVI and AQ/SAVI were treated with 400µM cleavable chemical crosslinker DSP (Pierce™, cat no: PG82081) in PBS for 1 hr at 4°C. Cells were washed with PBS and lysed in RIPA buffer. Whole cell lysate was mixed with 4x Laemmli Sample Buffer (BioRad, cat no 1610747) containing 5% 2-mercaptoethanol, heated at 95°C for 10 min and, run on a 4∼20% Mini-PROTEAN TGX Stain-Free Precast Gel. The blot was probed for STING antibody (Proteintech, 19851-1-AP). Densitometry was determined by ImageLab 5. Data are representative of three independent experiments. Graphs represent the mean with error bars indication s.e.m. p values are determined by unpaired student T-test (A-E) or one-way ANOVA Tukey’s multiple comparison test (D, E). * p<0.05, **p<0.01, ***p<0.001. n.s.: not significant; n.d: not detected.

The HAQ allele increased, and the AQ allele restored T-regs in SAVI(N153S) mice

IFNγ was proposed to drive SAVI disease 15,35,48. We confirmed that WT/SAVI CD4 T cells were enriched with IFNγ+ cells (Figure 6A). However, WT/SAVI mice have CD4 T cellpenia. Thus, the total numbers of spleen IFNγ+ CD4 T cells were comparable in WT/SAVI and AQ/SAVI mice (Figure 6A). In contrast, the HAQ/SAVI mice had decreased IFNγ+ CD4 T cells (Figure 6A).

HAQ/SAVI(N153S) and AQ/SAVI(N153S) cells had 10-fold and 20-fold increased spleen T-regs compared to WT/SAVI mice.

(A). Flow cytometry analysis of IFNγ producing spleen CD4+ T cells from WT/HAQ, WT/AQ, WT/SAVI, HAQ/SAVI and AQ/SAVI mice. (B). Flow cytometry analysis of CD4+ FoxP3+ spleen T-regs. n=3∼5 mice/group. Data are representative of 3 independent experiments. Graphs represent the mean with error bars indication s.e.m. p values are determined by one-way ANOVA Tukey’s multiple comparison test. * p<0.05, **p<0.01, ***p<0.001. n.s.: not significant.

The induction of Foxp3 expression in T-reg cells during ongoing autoimmune inflammation resolved inflammation and pathology in mice 49. CD4 T cellpenia depletes CD4 T-regs. Indeed, WT/SAVI mice had ∼20-fold reduction of spleen FoxP3+ T-regs compared to AQ/SAVI or WT/WT littermates (Figure 6B). The HAQ/SAVI mice also had ∼10-fold more T-regs than the WT/SAVI littermate (Figure 6B).

Discussion

This study, using the HAQ, AQ, SAVI(N153S) TMEM173 knock-in mice, reveals the in vivo significance and mechanism of STING-mediated CD4 T cell death. HAQ, AQ alleles prevent CD4 T cellpenia, and increase/restore CD4 T-regs in SAVI mice. The results are consistent with previous finding that the impaired CD4 T cell proliferation by the SAVI(V155M) mutant could be rescued by the addition of the HAQ allele in vitro 24. STING has been increasingly implicated in inflammatory diseases such as nonalcoholic fatty liver disease, nonalcoholic steatohepatitis, cardiomyopathy, obesity, diabetes, neurodegenerative diseases, aging, and kidney injury, many of which are independent of type I IFNs 2,3,9. It is tempting to suggest that STING activation in CD4 T cells leads to CD4 T-regs depletion that break tissue tolerance and exacerbates tissue inflammation.

Human immunodeficiency virus (HIV) primarily infects CD4 T cells and might activate the STING pathway in CD4 T cells 50–55. The loss of CD4 T cells is the hallmark of untreated HIV infection 56,57, and the measurement of CD4 T cell count is a central part of HIV care. We found that HAQ and AQ CD4 T cells are resistant to STING-mediated cell death. Mogensen and colleagues reported that HAQ/HAQ was enriched in HIV-infected long-term nonprogressors 39. These HAQ/HAQ individuals had reduced inhibition of CD4 T cell proliferation and a reduced immune response to DNA and HIV 39. It is likely that HIV infection activates STING-cell death pathway in CD4 T cells. In HAQ/SAVI and AQ/SAVI mice, one copy of HAQ, AQ allele suppressed CD4 T cell death. HAQ, AQ carriers might have fewer HIV-induced CD4 T cell death, thus being long-term nonprogressors in HIV infection-induced Acquired immunodeficiency syndrome (AIDS) 39. Targeting STING to prevent CD4 T cell death might be a valid therapy for AIDS. Activating the STING pathway is a promising strategy for cancer immunotherapy 58–64.

Multiple STING agonists are in clinical trials 44,45. Recently, the safety issue emerged in some STING agonist trials 44,45. For example, in the STING agonist, ADU-S100, clinical trial, Grade 3/4 treatment-related adverse events were reported in 12.2% of 41 pretreated patients (NCT02675439)44,45. The National Institutes of Health defines grade 3 as “incapacitating; unable to perform usual activities; requires absenteeism or bed rest.” In a clinical trial using STING antibody-drug-conjugate (ADC) that conjugates diABZI to anti-HER2 Ab, a Grade 5 (fatal) serious adverse event was recorded and deemed related to the STING-ADC (NCT05514717). SAVI disease, driven by overreacting STING, is often fatal. AQ, to a less degree, HAQ, suppress mortality in SAVI mice. Future STING clinical trials should be based on human TMEM173 genotype to achieve safe and effective responses.

Mechanistically, apoptosis, pyroptosis, ferroptosis, necroptosis, and PANoptosis have all been reported in STING-mediated cell death 25–31,42,43. Different cell types and STING agonists used likely contributed to the inconsistency and complexity. Here, we focused on lymphopenia in the SAVI mice that avoids ligand-dependent, non-physiological dosage in STING-mediated cell death. HAQ and AQ alleles could prevent CD4 T cellpenia in the SAVI mice strongly indicating that residue A230 or Q293 prevent STING-mediated CD4 T cell death in vivo. Splenocyte from Q293 mice were resistant to STING agonists-induced cell death ex vivo. Thus, it is likely that the Q293 residue is critical for STING-mediate lymphopenia. Notably, Q293 is outside the C-terminal tail (CTT) (residues 341–379 of human STING) critical for TBK1 recruitment and IRF3 phosphorylation 65 or miniCTT domain (aa343–354) 24, or the UPR motif (aa322–343) 46 important for T cells death in vitro. Further studies are needed to understand how the aa293 of STING mediates cell death in vivo. Noteworthy, AQ/SAVI cells had similar TBK1-IRF3, NFκB activation and STING degradation as the WT/SAVI cells. Yet, AQ/SAVI mice did not have CD4 T cellpenia as WT/SAVI mice suggesting that the canonical STING-TBK1-IRF3/NFκB pathway, likely STING oligomerization, is not sufficient for the induction of cell death at the physiological condition.

We used the WT/N153S knock-in SAVI mouse model that spontaneously develop lung inflammation, T cell cytopenia, and early mortality, mimicking pathological findings in human SAVI patients16. Using the WT/N153S SAVI mouse model and human Jurkat T cell line, it was proposed that STING activation causes chronic ER stress and unfolded protein response, leading to T cell death by apoptosis 46. Furthermore, the study showed that crossing WT/N153S mice to the OT-I mice reduced ER stress and restored CD8+, but not CD4+, T cells 46. The restoration of CD8+T cells reduces inflammation and lung disease 46. However, human WT/N154S SAVI patients have normal CD8+ T cells numbers 13, and primary human CD8+ T cells are largely resistant to STING-agonists-induced cell death ex vivo (Figure 2A) 25. Thus, it is puzzling how restoring CD8+ T cells can rescue SAVI phenotypes since the SAVI patients already have normal CD8+ T cells numbers.

Finally, it is unexpected that both HAQ and AQ alleles are resistant to cell death. Our previous studies showed that the HAQ and AQ alleles have opposite functions 34. AQ-STING, not HAQ-STING, responds to CDNs 32–34,36–40,66. AQ mice are lean while HAQ mice are fat 34. Most importantly, HAQ was positively selected, while AQ was negatively selected, in modern humans outside Africans 34. Thus, the death pathway of STING is also distinct from the STING function that was naturally selected. Together, we propose that there are at least three unique pathways of STING/MPYS: STING-Type I IFNs, STING-cell death, and STING-fatty acid metabolism.

The limitations of the study

The poor transferability of mouse to humans is a major issue in STING research 44,45. The present study used AQ/SAVI and HAQ/SAVI mice. Confirmation is needed in humans with the identification and evaluation of people who are AQ/SAVI, HAQ/SAVI.

Materials and Methods

Experimental Design

The study was designed to reveal (i) the in vivo significance of the type I IFNs-independent, STING-dependent cell death function; (ii) the interplay between common STING alleles HAQ, AQ and the rare, gain-of-function SAVI STING mutation; (iii). The driver for the inflammatory SAVI disease. Mouse splenocytes, primary human lung cells, human THP-1 cells and HAQ, AQ, SAVI knock-in mice were used to establish the in vivo significance and human relevance. All the repeats were biological replications that involve the same experimental procedures on different mice. Where possible, treatments were assigned blindly to the experimenter by another individual in the lab. When comparing samples from different groups, samples from each group were analyzed in concert, thereby preventing any biases that might arise from analyzing individual treatments on different days. All experiments were repeated at least twice.

Mice

WT/SAVI(N153S) mice were purchased from The Jackson Laboratory. HAQ, AQ mice were previously generated in the lab 33,34. The Q293 mice were generated by Cyagen Biosciences. Briefly, the linearized targeting vector was transfected into JM8A3.N1 C57BL/6N embryonic stem cells. A positive embryonic stem clone was subjected to the generation of chimera mice by injection using C57BL/6J blastocysts as the host. Successful germline transmission was confirmed by PCR sequencing. The heterozygous mice were bred to Actin-flpase mice [The Jackson Laboratory, B6.Cg-Tg (ACTFLPe)9205Dym/J] to remove the neo gene and make the Q293 knock-in mouse. Age- and gender-matched mice (2 – 6 month old, both male and female) were used for indicated experiments. WT/SAVI (male) x WT/HAQ (female), WT/SAVI (male) x WT/AQ (female) breeders were set up to generate HAQ/SAVI, AQ/SAVI mice. Mice were housed at 22°C under a 12-h light-dark cycle with ad libitum access to water and a chow diet (3.1 kcal/g, Teklad 2018, Envigo, Sommerset, NJ) and bred under pathogen-free conditions in the Animal Research Facility at the University of Florida. Littermates of the same sex were randomly assigned to experimental groups. All mouse experiments were performed by the regulations and approval of the Institutional Animal Care and Use Committee at the University of Florida, IACUC202200000058.

Reagents

Recombinant human IFNβ (R&D, cat no. 8499-IF-010/CF), diABZI (Invivogen, cat no. 2138299-34-8), 2’3’-cGAMP (Invivogen, cat no. tlrl-nacga23-02), DMXAA (Invivogen, cat no. tlrl-dmx), H151 (Invivogen, cat no. inh-h151), RpRpSSCDA (Biolog, cat no. C118), THP1-Dual™ KO-STING Cells (Invivogen, cat no. thpd-kostg). All other chemical inhibitors are from Selleckchem. Mouse TNF alpha ELISA Ready Set Go. (eBioscience, cat no. 88-7324). Mouse IFN-Beta ELISA Kit (PBI, cat no. 42400).

Histology

Lungs and livers were fixed in 10% formalin, paraffin-embedded, and cut into 4-µm sections. Lung, liver sections were then stained for hematoxylin-eosin. All staining procedures were performed by the histology core at the University of Florida. Briefly, tissue sectins were immersed Harris Hematoxylin for 10 seconds, then washed with tap water. Cleard sections were re-immersed in EOSIN stain for ∼30 seconds. The sections were washed with tap water until clear, then dehydrate in ascending alcohol solutions (50%,70%,80%,95% x 2, 100% x 2). Afterwards, the sections werer cleared with xylene (3 - 4 x). The sections were mounted on glass slide with permount organic mounting medium for visulization.

Lung Function

Pulmonary function was evaluated using an isolated, buffer-perfused mouse lung apparatus (Hugo Sachs Elektronik, March-Huggstetten, Germany), as previously described 67. Briefly, mice were anesthetized with ketamine and xylazine and a tracheostomy was performed, and animals were ventilated with room air at 100 breaths/min at a tidal volume of 7 μl/g body weight with a positive end-expiratory pressure of 2 cm H2O using a pressure-controlled ventilator (Hugo Sachs Elektronik, March-Huggstetten, Germany).

Isolation of lung cells

Cells were isolated from the lung as previously described 68. The lungs were perfused with ice-cold PBS and removed. Lungs were digested in DMEM containing 200μg/ml DNase I (Roche, 10104159001), 25μg/ml Liberase TM (Roche, 05401119001) at 37°C for 2 hours. Red blood cells were then lysed and a single cell suspension was prepared by filtering through a 70-µm cell strainer.

BMDM activation

BMDMs were induced from mouse bone marrow cells cultured in RPMI 1640 (cat#11965; Invitrogen) with 10% FBS, 2 mM L-glutamine, 1 mM sodium pyruvate, 10 mM HEPES buffer, 1% nonessential amino acids, 50 mM 2-ME, 1% Pen/Strep, with 20ng/ml M-CSF (Kingfisher, RP0407M) for 10 days 33. STING agonists were added into the culture (no transfection or membrane permeabilization).

Flow cytometry

Single cell suspensions were stained with fluorescent-dye-conjugated antibodies in PBS containing 2% FBS and 1mM EDTA. Surface stains were performed at 4°C for 20 min. For intracellular cytokine or transcription factor staining of murine and human cells, cells were fixed and permeabilized with the Foxp3 staining buffer set (eBioscience, cat no 00-5523-00). Cells were washed and stained with surface markers. Cells were then fixed and permeabilized (eBioscience, cat no. 00-5523-00) for intracellular cytokine stain. Data were acquired on a BD LSRFortessa and analyzed using the FlowJo software package (FlowJo, LLC). Cell sorting was performed on the BD FACSAriaIII Flow Cytometer and Cell Sorter. More information on the antibodies used can be found in supplementary table S1.

Human lung explants

Human lung explants were procured at the Lung Transplant Center, Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University of Florida. Donor and patients consent was obtained for a research protocol (UF IRB201902955-Treatment with IFNβ Induces Tolerogenic Lung Dendritic Cells in Human advanced lung disease). Healthy donor lungs were surgically removed postmortem, perfused, small pieces were cut from the right middle and lower lobes for research purpose, and stored in cold Perfadex ® at 4°C for no more than 12 hrs before processing. Ex planted lungs from emphysema lung transplant patients were stored in cold Perfadex ® at 4°C for no more than 12 hrs before the process. No lung explants were procured from prisoners.

Statistical Analysis

To gain statistical power, we employ three∼four mice/groups to characterize lung immunity. Ten mice/group to monitor animal health. The statistical justification for group size was calculated using the SAS program to calculate the animal numbers. The analysis was carried out using a standard error of 0.5 for immunological assays, and a power of 0.9. All data are expressed as means ± SEM. Statistical significance was evaluated using Prism 9.0 software. Comparisons between two groups were analyzed by performing an unpaired Student’s t test. Comparisons between more than two groups were analyzed by performing a one-way analysis of variance (ANOVA) with Tukey’s multiple comparisons test.

eLife Assessment

This study describes useful mouse models of knock-ins of human STING1 variants and an assessment of these variants' action in mouse immune cells. While the data included in the manuscript are solid, because of the authors' interpretation and conclusions made, the work is currently considered incomplete.

Reviewer #1 (Public Review):

Summary:

This manuscript by Aybar-Torres et al investigated the effect of common human STING1 variants on STING-mediated T cell phenotypes in mice. The authors previously made knock-in mice expressing human STING1 alleles HAQ or AQ, and here they established a new knock-in line Q293. The authors stimulated cells isolated from these mice with STING agonists and found that all three human mutant alleles resist cell death, leading to the conclusion that R293 residue is essential for STING-mediated cell death (there are several caveats with this conclusion, more below). The authors also bred HAQ and AQ alleles to the mouse Sting1-N153S SAVI mouse and observed varying levels of rescue of disease phenotypes with the AQ allele showing more complete rescue than the HAQ allele. The Q293 allele was not tested in the SAVI model. They conclude that the human common variants such as HAQ and AQ have a dominant negative effect over the gain-of-function SAVI mutants.

Strengths:

The authors and Dr. Jin's group previously made important observations of common human STING1 variants, and these knock-in mouse models are essential for understanding the physiological function of these alleles.

Weaknesses:

However, although some of the observations reported here are interesting, the data collectively does not support a unified model. The authors seem to be drawing two sets of conclusions from in vitro and in vivo experiments, and neither mechanism is clear. Several experiments need better controls, and these knock-in mice need more comprehensive functional characterization.

(1) In Figure 1, the authors are trying to show that STING agonist-induced splenocytes cell death is blocked by HAQ, AQ and Q alleles. The conclusion at line 134 should be splenocytes, not lymphocytes. Most experiments in this figure were done with mixed population that may involve cell-to-cell communication. Although TBK1-dependence is likely, a single inhibitor treatment of a mixed population is not sufficient to reach this conclusion.

(2) Q293 knock-in mouse needs to be characterized and compared to HAQ and AQ. Is this mutant expressed in tissues? Does this mutant still produce IFN and other STING activities? Does the protein expression level altered on Western blot? Is the mutant protein trafficking affected? In the authors' previous publications and some of the Western blot here, expression levels of each of these human STING1 protein in mice are drastically different. HAQ and AQ also have different effects on metabolism (pmid: 36261171), which could complicate interoperation of the T cell phenotypes.

(3) HAQ/WT and AQ/WT splenocytes are protected from STING agonist-induced cell death equally well (Figure 1G). HAQ/SAVI shows less rescue compared to AQ/SAVI. These are interesting observations, but mechanism is unclear and not clearly discussed. E.g., how does AQ protect disease pathology better than HAQ (that contains AQ)? Does Q293 allele also fully rescue SAVI?

(4) Figure 2 feels out of place. First of all, why are the authors using human explant lung tissues? PBMCs should be a better source for lymphocytes. In untreated conditions, both CD4 and B cells show ~30% dying cells, but CD8 cells show 0% dying cells. This calls for technical concerns on the CD8 T cell property or gating strategy because in the mouse experiment (Figure 1A) all primary lymphocytes show ~30% cell death at steady-state. Second, Figure 2C, these type of partial effect needs multiple human donors to confirm. Three, the reconstitution of THP1 cells seems out of place. STING-mediated cell death mechanism in myeloid and lymphoid cells are likely different. If the authors want to demonstrate cell death in myeloid cells using THP1, then these reconstituted cell lines need to be better validated. Expression, IFN signaling, etc. The parental THP1 cells is HAQ/HAQ, how does that compare to the reconstitutions? There are published studies showing THP1-STING-KO cells reconstituted with human variants do not respond to STING agonists as expected. The authors need to be scientifically rigorous on validation and caution on their interpretations.

(5) Figure 2G, H, I are confusing. AQ is more active in producing IFN signaling than HAQ and Q is the least active. How to explain this?

(6) The overall model is unclear. If HAQ, AQ and Q are loss-of-function alleles and Q is the key residue for STING-mediated cell death, then why AQ is the most active in producing IFN signaling and AQ/SAVI rescues disease most completely? If these human variants act as dominant negatives, which would be consistent with the WT/het data, then how do you explain AQ is more dominant negative than HAQ?

(7) As a general note, SAVI disease phenotypes involve multiple cell types. Lymphocyte cell death is only one of them. The authors' characterization of SAVI pathology is limited and did not analyze immunopathology of the lung.

(8) Line 281, the discussion on HIV T cell death mechanism is not relevant and over-stretching. This study did not evaluate viral infection in T cells at all. The original finding of HAQ/HAQ enrichment in HIV/AIDS was 2/11 in LTNP vs 0/11 in control, arguably not the strongest statistics.

Reviewer #2 (Public Review):

Aybar-Torres and colleagues utilize common human STING alleles to dissect the mechanism of SAVI inflammatory disease. The authors demonstrate that these common alleles alleviate SAVI pathology in mice, and perhaps more importantly use the differing functionality of these alleles to provide insight into requirements of SAVI disease induction. Their findings suggest that it is residue A230 and/or Q293 that are required for SAVI induction, while the ability to induce an interferon-dependent inflammatory response is not. This is nicely exemplified by the AQ/SAVI mice that have an intact inflammatory response to STING activation, yet minimal disease progression. As both mutants seem to be resistant STING-dependent cell death, this manuscript also alludes to the importance of STING-dependent cell death, rather than STING-dependent inflammation, in the progression of SAVI pathology. While I have some concerns, I believe this manuscript makes some important connections between STING pathology mouse models and human genetics that would contribute to the field.

Some points to consider:

(1) While the CD4+ T cell counts from HAQ/SAVI and AQ/SAVI mice suggest that these T cells are protected from STING-dependent cell death, an assay that explores this more directly would strengthen the manuscript. This is also supported by Fig 2C, but I believe a strength of this manuscript is the comparison between the two alleles. Therefore, if possible, I would recommend the isolation of T cells from these mice and direct stimulation with diABZI or other STING agonist with a cell death readout.

(2) Related to the above point - further exemplifying that the Q293 locus is essential to disease, even in human cells, would also strengthen the paper. It seems that CD4 T cell loss is a major component of human SAVI. While not completely necessary, repeating the THP1 cell death experiments from Fig 2 with a human T cell line would round out the study nicely.

(3) While I found the myeloid cell counts and BMDM data interesting, I think some more context is needed to fully loop this data into the story. Is myeloid cell expansion exemplified by SAVI patients? Do we know if myeloid cells are the major contributors to the inflammation these patients experience? Why should the SAVI community care about the Q293 locus in myeloid cells?

(4) The functional assays in Figure 4 are exciting and really connect the alleles to disease progression. To strengthen the manuscript and connect all the data, I would recommend additional readouts from these mice that address the inflammatory phenotype shown in vitro in Figure 5. For example, measuring cytokines from these mice via ELISA or perhaps even Western blots looking for NFkB or STING activation would be supportive of the story. This would also allow for some tissue specificity. I believe looking for evidence of inflammation and STING activation in the lungs of these mice, for example, would further connect the data to human SAVI pathology.

Author response:

The following is the authors’ response to the original reviews.

Summary Responses: Besides the WT allele, equivalent to the mouse TMEM173 gene, the human TMEM173 gene has two common alleles: the HAQ and AQ alleles carried by billions of people. The main conclusions and interpretation, summarized in the Title and Abstract, are i) Different from the WT TMEM173 allele, the HAQ or AQ alleles are resistant to STING activation-induced cell death; ii) STING residue 293 is critical for cell death; iii) HAQ, AQ alleles are dominant to the SAVI allele; iv) One copy of the AQ allele rescues the SAVI disease in mice. We propose that STING research and STING-targeting immunotherapy should consider human TMEM173 heterogeneity. These interpretations and conclusions were based on Data and Logic. We welcome alternative, logical interpretations and collaborations to advance the human TMEM173 research.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

This manuscript by Aybar-Torres et al investigated the effect of common human STING1 variants on STING-mediated T cell phenotypes in mice. The authors previously made knock-in mice expressing human STING1 alleles HAQ or AQ, and here they established a new knock-in line Q293. The authors stimulated cells isolated from these mice with STING agonists and found that all three human mutant alleles resist cell death, leading to the conclusion that R293 residue is essential for STING-mediated cell death (there are several caveats with this conclusion, more below). The authors also bred HAQ and AQ alleles to the mouse Sting1-N153S SAVI mouse and observed varying levels of rescue of disease phenotypes with the AQ allele showing more complete rescue than the HAQ allele. The Q293 allele was not tested in the SAVI model. They conclude that the human common variants such as HAQ and AQ have a dominant negative effect over the gain-of-function SAVI mutants.

Strengths:

The authors and Dr. Jin's group previously made important observations of common human STING1 variants, and these knock-in mouse models are essential for understanding the physiological function of these alleles.

Weaknesses:

However, although some of the observations reported here are interesting, the data collectively does not support a unified model. The authors seem to be drawing two sets of conclusions from in vitro and in vivo experiments, and neither mechanism is clear. Several experiments need better controls, and these knock-in mice need more comprehensive functional characterization.

(1) In Figure 1, the authors are trying to show that STING agonist-induced splenocytes cell death is blocked by HAQ, AQ and Q alleles. The conclusion at line 134 should be splenocytes, not lymphocytes. Most experiments in this figure were done with mixed population that may involve cell-to-cell communication. Although TBK1-dependence is likely, a single inhibitor treatment of a mixed population is not sufficient to reach this conclusion.

We greatly appreciate Reviewer 1's insights. We changed the “lymphocytes” to “splenocytes” (line 133) as suggested. We respectfully disagree with Reviewer 1’s comments on TBK1. First, we used two different TBK1 inhibitors: BX795 and GSK8612. Second, because BX795 also inhibits PDK1, we used a PDK1 inhibitor GSK2334470; Third, both BX795 and GSK8612 completely inhibited diABZI-induced splenocyte cell death (Figure 1B) (lines 128 – 133). The logical conclusion is “TBK1 activation is required for STING-mediated mouse spleen cell death ex vivo”. (line 117).

Our discovery that the common human TMEM173 alleles are resistant to STING activation-induced cell death is a substantial finding. It further strengthens the argument that the HAQ and AQ alleles are functionally distinct from the WT allele 1-3. We wish to underscore the crucial message of this study-that 'STING research and STING-targeting immunotherapy should consider TMEM173 heterogeneity in humans' (line 37), which has been largely overlooked in current STING clinical trials 4.

Regarding STING-Cell death, as we stated in the Introduction (lines 65-77). i) STING-mediated cell death is cell type-dependent 5-7 and type I IFNs-independent 5,7,8. ii) The in vivo biological significance of STING-mediated cell death is not clear 7,8. iii) The mechanisms of STING-Cell death remain controversial. Multiple cell death pathways, i.e., apoptosis, necroptosis, pyroptosis, ferroptosis, and PANoptosis, are proposed 7,9,10. SAVI/HAQ, SAVI/AQ prevented lymphopenia and alleviated SAVI disease in mice. Thus, the manuscript provides some answers to the biological significance of STING-cell death in vivo, which is new. Regarding the molecular mechanism, splenocytes from Q293/Q293 mice are resistant to STING cell death. The logical conclusion is that the amino acid 293 is critical for STING cell death (line 29).

Extensive studies are needed, beyond the scope of this manuscript, on how aa293 and TBK1 mediates STING-Cell death to resolve the controversies in the STING-cell death fields (e.g. apoptosis, necroptosis, pyroptosis, ferroptosis, and PANoptosis).

(2) Q293 knock-in mouse needs to be characterized and compared to HAQ and AQ. Is this mutant expressed in tissues? Does this mutant still produce IFN and other STING activities? Does the protein expression level altered on Western blot? Is the mutant protein trafficking affected? In the authors' previous publications and some of the Western blot here, expression levels of each of these human STING1 protein in mice are drastically different. HAQ and AQ also have different effects on metabolism (pmid: 36261171), which could complicate interoperation of the T cell phenotypes.

These are very important questions that require rigorous investigations that are beyond the scope of this manuscript. This manuscript, titled “The common TMEM173 HAQ, AQ alleles rescue CD4 T cellpenia, restore T-regs, and prevent SAVI (N153S) inflammatory disease in mice” does not focus on Q293 mice. We have been investigating the common human TMEM173 alleles since 2011 from the discovery 11 , mouse model 1,3, human clinical trial 2, and human genetics studies 3. This manuscript is another step towards understanding these common human TMEM173 alleles with the new discovery that HAQ, AQ alleles are resistant to STING cell death.

(3) HAQ/WT and AQ/WT splenocytes are protected from STING agonist-induced cell death equally well (Figure 1G). HAQ/SAVI shows less rescue compared to AQ/SAVI. These are interesting observations, but mechanism is unclear and not clearly discussed. E.g., how does AQ protect disease pathology better than HAQ (that contains AQ)? Does Q293 allele also fully rescue SAVI?

In this manuscript, Figure 6 shows AQ/SAVI had more T-regs than HAQ/SAVI (lines 251 – 261). In our previous publication on HAQ, AQ knockin mice, we showed that AQ T-regs have more IL-10 than HAQ T-regs 3. Thus, increased IL-10+ Tregs in AQ mice may contribute to an improved phenotype in AQ/SAVI compared to HAQ/SAVI. However, we are not excluding other contributions (e.g. metabolic difference) (lines 332-335). We are exploring these possibilities.

(4) Figure 2 feels out of place. First of all, why are the authors using human explant lung tissues? PBMCs should be a better source for lymphocytes. In untreated conditions, both CD4 and B cells show ~30% dying cells, but CD8 cells show 0% dying cells. This calls for technical concerns on the CD8 T cell property or gating strategy because in the mouse experiment (Figure 1A) all primary lymphocytes show ~30% cell death at steady-state. Second, Figure 2C, these type of partial effect needs multiple human donors to confirm. Three, the reconstitution of THP1 cells seems out of place. STING-mediated cell death mechanism in myeloid and lymphoid cells are likely different. If the authors want to demonstrate cell death in myeloid cells using THP1, then these reconstituted cell lines need to be better validated. Expression, IFN signaling, etc. The parental THP1 cells is HAQ/HAQ, how does that compare to the reconstitutions? There are published studies showing THP1-STING-KO cells reconstituted with human variants do not respond to STING agonists as expected. The authors need to be scientifically rigorous on validation and caution on their interpretations.

Figure 2 is necessary because it reveals the difference between mouse and human STING cell death, which is critical to understand STING in human health and diseases (lines 160-161). Figure 2A-2B showed that STING activation killed human CD4 T cells, but not human CD8 T cells or B cells. This observation is different from Figure 1A, where STING activation killed mouse CD4, CD8 T cells, and CD19 B cells, revealing the species-specific STING cell death responses. Regarding human CD8 T cells, as we stated in the Discussion (lines 323-325), human CD8 T cells (PBMC) are not as susceptible as the CD4 T cells to STING-induced cell death 8. We used lung lymphocytes that showed similar observations (Figure 2A). For Figure 2C, we used 2 WT/HAQ and 3 WT/WT individuals (lines 738-739). We generate HAQ, AQ THP-1 cells in STING-KO THP-1 cells (Invivogen,, cat no. thpd-kostg) (lines 380-387).

A recent study found that a new STING agonist SHR1032 induces cell death in STING-KO THP-1 cells expressing WT(R232) human STING 10 (line 182). SHR1032 suppressed THP1-STING-WT(R232) cell growth at GI50: 23 nM while in the parental THP1-STING-HAQ cells, the GI50 of SHR1032 was >103 nM 10. Cytarabine was used as an internal control where SHR1032 killed more robustly than cytarabine in the THP1-STING-WT(R232) cells but much less efficiently than cytarabine in the THP-1-STING-HAQ cells 10.

Our manuscript rigorously uses mouse splenocytes, human lung lymphocytes, THP-1 reconstituted with HAQ, AQ, and HAQ/SAVI, AQ/SAVI mice, to demonstrate that the common human HAQ, AQ alleles are resistant to STING cell death in vitro and in vivo.

We agree with Reviewer 1 that STING-mediated cell death mechanisms in myeloid and lymphoid cells may be different and likely contribute to the different mechanisms proposed in STING cell death research 7,9,10. Our study focuses on the in vivo STING-mediated T cellpenia.

(5) Figure 2G, H, I are confusing. AQ is more active in producing IFN signaling than HAQ and Q is the least active. How to explain this?

We stated in the Introduction that “AQ responds to CDNs and produce type I IFNs in vivo and in vitro 3,12,13 ”(line 92-93). We reported that the AQ knock in mice responded to STING activation 3. We previously showed that there was a negative natural selection on the AQ allele in individuals outside of Africa 3. 28% of Africans are WT/AQ but only 0.6% East Asians are WT/AQ 3. In contrast, the HAQ allele was positively selected in non-Africans 3. Investigation to understand the mechanisms and biological significance of these naturally selected human TMEM173 alleles has been ongoing in the lab.

(6) The overall model is unclear. If HAQ, AQ and Q are loss-of-function alleles and Q is the key residue for STING-mediated cell death, then why AQ is the most active in producing IFN signaling and AQ/SAVI rescues disease most completely? If these human variants act as dominant negatives, which would be consistent with the WT/het data, then how do you explain AQ is more dominant negative than HAQ?

In this manuscript, Figure 6 shows AQ/SAVI had more T-regs than HAQ/SAVI (lines 251 – 261). In our previous publication on HAQ, AQ knockin mice, we showed that AQ T-regs have more IL-10 and mitochondria activity than HAQ T-regs 3. Nevertheless, we are not excluding other contributions (e.g. metabolic difference) by the AQ allele (lines 332-335). Last, we used modern human evolution to discover the dominance of these common human STING alleles. In modern humans outside Africans, HAQ was positively selected while AQ was negatively selected 3. However, AQ is likely dominant to HAQ because there is no HAQ/AQ individuals outside Africa. The genetic dominance of common human TMEM173 allele is a new concept. More investigation is ongoing.

(7) As a general note, SAVI disease phenotypes involve multiple cell types. Lymphocyte cell death is only one of them. The authors' characterization of SAVI pathology is limited and did not analyze immunopathology of the lung.

Both radioresistant parenchymal and/or stromal cells and hematopoietic cells influence SAVI pathology in mice 14,15. Nevertheless, the lack of CD 4 T cells, including the anti-inflammatory T-regs, likely contributes to the inflammation in SAVI mice and patients 16. We characterized lung function, lung inflammation (Figure 4), lung neutrophils, and inflammatory monocyte infiltration (Figure S5) (lines 232-235).

(8) Line 281, the discussion on HIV T cell death mechanism is not relevant and over-stretching. This study did not evaluate viral infection in T cells at all. The original finding of HAQ/HAQ enrichment in HIV/AIDS was 2/11 in LTNP vs 0/11 in control, arguably not the strongest statistics.

Several publications have linked STING to HIV pathogenesis 17-22 (line 271). CD4 T cellpenia is a hallmark of AIDS. The manuscript studies STING activation-induced T cellpenia in vivo. It is not stretching to ask, for example, does preventing STING T cell death (e.g HAQ, AQ alleles) can restore CD4 T cell counts and improve care for AIDS patients?

Reviewer #2 (Public Review):

Aybar-Torres and colleagues utilize common human STING alleles to dissect the mechanism of SAVI inflammatory disease. The authors demonstrate that these common alleles alleviate SAVI pathology in mice, and perhaps more importantly use the differing functionality of these alleles to provide insight into requirements of SAVI disease induction. Their findings suggest that it is residue A230 and/or Q293 that are required for SAVI induction, while the ability to induce an interferon-dependent inflammatory response is not. This is nicely exemplified by the AQ/SAVI mice that have an intact inflammatory response to STING activation, yet minimal disease progression. As both mutants seem to be resistant STING-dependent cell death, this manuscript also alludes to the importance of STING-dependent cell death, rather than STING-dependent inflammation, in the progression of SAVI pathology. While I have some concerns, I believe this manuscript makes some important connections between STING pathology mouse models and human genetics that would contribute to the field.

Some points to consider:

(1) While the CD4+ T cell counts from HAQ/SAVI and AQ/SAVI mice suggest that these T cells are protected from STING-dependent cell death, an assay that explores this more directly would strengthen the manuscript. This is also supported by Fig 2C, but I believe a strength of this manuscript is the comparison between the two alleles. Therefore, if possible, I would recommend the isolation of T cells from these mice and direct stimulation with diABZI or other STING agonist with a cell death readout.

Please see the new Figure S3 for cell death by diABZI, DMXAA in Splenocytes from WT/WT, WT/HAQ, HAQ/SAVI, AQ/SAVI mice. The HAQ/SAVI and AQ/SAVI splenocytes showed similar partial resistance to STING activation-induced cell death (lines 214-216).

(2) Related to the above point - further exemplifying that the Q293 locus is essential to disease, even in human cells, would also strengthen the paper. It seems that CD4 T cell loss is a major component of human SAVI. While not co_mpletely necessary, repeating the THP1 cell death experiments from Fig 2 with a human T cell line would round out the study nicely._

We examined HAQ, AQ mouse splenocytes, HAQ human lung lymphocytes, THP-1 reconstituted with HAQ, AQ, and HAQ/SAVI, AQ/SAVI mice, to demonstrate that the common human HAQ, AQ alleles are resistant to STING cell death in vitro and in vivo. Additional human T cell line work does not add too much. We hope to conduct more human PBMC or lung lymphocytes STING cell death experiments from HAQ, AQ individuals as we continue the human STING alleles investigation.

(3) While I found the myeloid cell counts and BMDM data interesting, I think some more context is needed to fully loop this data into the story. Is myeloid cell expansion exemplified by SAVI patients? Do we know if myeloid cells are the major contributors to the inflammation these patients experience? Why should the SAVI community care about the Q293 locus in myeloid cells?

This is likely a misunderstanding. We use BMDM for the purpose of comparing STING signaling (TBK1, IRF3, NFkB, STING activation) by WT/SAVI, HAQ/SAVI, AQ/SAVI. Ideally, we would like to compare STING signaling in CD4 T cells from WT/SAVI to HAQ/SAVI, AQ/SAVI mice. However, WT/SAVI has no CD4 T cells. Doing so, we are making the assumption that the basic STING signaling (TBK1, IRF3, NFkB, STING activation) is conserved between T cells and macrophages.

(4) The functional assays in Figure 4 are exciting and really connect the alleles to disease progression. To strengthen the manuscript and connect all the data, I would recommend additional readouts from these mice that address the inflammatory phenotype shown in vitro in Figure 5. For example, measuring cytokines from these mice via ELISA or perhaps even Western blots looking for NFkB or STING activation would be supportive of the story. This would also allow for some tissue specificity. I believe looking for evidence of inflammation and STING activation in the lungs of these mice, for example, would further connect the data to human SAVI pathology.

Reviewer 2 suggests looking for evidence of inflammation and STING activation in the lungs of HAQ/SAVI, AQ/SAVI. We would like to elaborate further. First, anti-inflammatory treatments, e.g. steroids, DMARDs, IVIG, Etanercept (TNF), rituximab, Nifedipine, amlodipine, et al., all failed in SAVI patients 23. JAK inhibitors on SAVI had mixed outcomes (lines 55-58). Second, Figure S5 examined lung neutrophils and inflammatory monocyte infiltration. Interestingly, while AQ/SAVI mice had a better lung function than HAQ/SAVI mice (Figure 4D, 4E vs 4H, 4I), HAQ/SAVI and AQ/SAVI lungs had comparable neutrophils and inflammatory monocyte infiltration (Figure S5). Last, SAVI is classified as type I interferonopathy 23, but the lung diseases of SAVI are mainly independent of type I IFNs 24-27. The AQ allele suppresses SAVI in vivo. Understanding the mechanisms by which AQ rescues SAVI may lead to curative care for SAVI patients.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

One suggestion is to streamline this study by focusing on STING-mediated cell death only in CD4 T cells. The authors can use in vitro PBMC isolated human T cells, ex vivo T cells from the knock-in mice, and in vivo T cells from the SAVI breeding. The current manuscript includes myeloid cell death, Tregs, complex SAVI disease pathology, which is too confusing and too complex to explain with the varying effect from the three human STING1 variants.

We sincerely appreciate Reviewer 1’s suggestion. The goal of our human STING alleles research has always been translational, i.e. improving human health. Even as a monogenetic disease, the SAVI pathology is still complex. For example, thought as a type I Interferonopathy, SAVI is largely independent of type I IFNs. Similarly, STING-activation-induced cell death, while contribute to SAVI, is not the whole story, as the Reviewer pointed out in the Comment 3 & 6 &7. HAQ/SAVI mice still died early and had lung dysfunction (Figure 4). In contrast, AQ/SAVI mice restore lifespan and lung function. We had Figure 6 show different T-regs between AQ/SAVI and HAQ/SAVI mice. In addition, AQ mice had more IL-10+ T-regs than HAQ mice 3. Therefore, we are excited about developing AQ-based curative therapy for SAVI patients (preventing cell death and inducing immune tolerance). Again, we thank the Reviewer for the suggestion. Additional research is ongoing.

Reviewer #2 (Recommendations For The Authors):

Minor points

(1) Generation of THP1 cells with the human STING alleles is missing from methods.

We added the protocol in the methods (lines 380-387). THP-1 KO line stable expressing WT STING was first described by Weikang Tao’s group 10.

(2) Some abbreviations are not expanded (CDA).

CDA is expanded as cyclic di-AMP (e.g. line 375).

References.

(1) Patel, S. et al. The Common R71H-G230A-R293Q Human TMEM173 Is a Null Allele. J Immunol 198, 776-787 (2017).

(2) Sebastian, M. et al. Obesity and STING1 genotype associate with 23-valent pneumococcal vaccination efficacy. JCI Insight 5 (2020).

(3) Mansouri, S. et al. MPYS Modulates Fatty Acid Metabolism and Immune Tolerance at Homeostasis Independent of Type I IFNs. J Immunol 209, 2114-2132 (2022).

(4) Sivick, K. E. et al. Comment on "The Common R71H-G230A-R293Q Human TMEM173 Is a Null Allele". J Immunol 198, 4183-4185 (2017).

(5) Gulen, M. F. et al. Signalling strength determines proapoptotic functions of STING. Nat Commun 8, 427 (2017).

(6) Kabelitz, D. et al. Signal strength of STING activation determines cytokine plasticity and cell death in human monocytes. Sci Rep 12, 17827 (2022).

(7) Murthy, A. M. V., Robinson, N. & Kumar, S. Crosstalk between cGAS-STING signaling and cell death. Cell Death Differ 27, 2989-3003 (2020).

(8) Kuhl, N. et al. STING agonism turns human T cells into interferon-producing cells but impedes their functionality. EMBO Rep 24, e55536 (2023).

(9) Li, C., Liu, J., Hou, W., Kang, R. & Tang, D. STING1 Promotes Ferroptosis Through MFN1/2-Dependent Mitochondrial Fusion. Front Cell Dev Biol 9, 698679 (2021).

(10) Song, C. et al. SHR1032, a novel STING agonist, stimulates anti-tumor immunity and directly induces AML apoptosis. Sci Rep 12, 8579 (2022).

(11) Jin, L. et al. Identification and characterization of a loss-of-function human MPYS variant. Genes Immun 12, 263-269 (2011).

(12) Yi, G. et al. Single nucleotide polymorphisms of human STING can affect innate immune response to cyclic dinucleotides. PLoS One 8, e77846 (2013).

(13) Patel, S. et al. Response to Comment on "The Common R71H-G230A-R293Q Human TMEM173 Is a Null Allele". J Immunol 198, 4185-4188 (2017).

(14) Gao, K. M. et al. Endothelial cell expression of a STING gain-of-function mutation initiates pulmonary lymphocytic infiltration. Cell Rep 43, 114114 (2024).

(15) Gao, K. M., Motwani, M., Tedder, T., Marshak-Rothstein, A. & Fitzgerald, K. A. Radioresistant cells initiate lymphocyte-dependent lung inflammation and IFNgamma-dependent mortality in STING gain-of-function mice. Proc Natl Acad Sci U S A 119, e2202327119 (2022).

(16) Hu, W. et al. Regulatory T cells function in established systemic inflammation and reverse fatal autoimmunity. Nat Immunol 22, 1163-1174 (2021).

(17) Monroe, K. M. et al. IFI16 DNA sensor is required for death of lymphoid CD4 T cells abortively infected with HIV. Science 343, 428-432 (2014).

(18) Doitsh, G. et al. Cell death by pyroptosis drives CD4 T-cell depletion in HIV-1 infection. Nature 505, 509-514 (2014).

(19) Jakobsen, M. R., Olagnier, D. & Hiscott, J. Innate immune sensing of HIV-1 infection. Curr Opin HIV AIDS 10, 96-102 (2015).

(20) Silvin, A. & Manel, N. Innate immune sensing of HIV infection. Curr Opin Immunol 32, 54-60 (2015).

(21) Altfeld, M. & Gale, M., Jr. Innate immunity against HIV-1 infection. Nat Immunol 16, 554-562 (2015).

(22) Krapp, C., Jonsson, K. & Jakobsen, M. R. STING dependent sensing - Does HIV actually care? Cytokine Growth Factor Rev 40, 68-76 (2018).

(23) Liu, Y. et al. Activated STING in a vascular and pulmonary syndrome. N Engl J Med 371, 507-518 (2014).

(24) Luksch, H. et al. STING-associated lung disease in mice relies on T cells but not type I interferon. J Allergy Clin Immunol 144, 254-266 e258 (2019).

(25) Stinson, W. A. et al. The IFN-gamma receptor promotes immune dysregulation and disease in STING gain-of-function mice. JCI Insight 7 (2022).

(26) Warner, J. D. et al. STING-associated vasculopathy develops independently of IRF3 in mice. J Exp Med 214, 3279-3292 (2017).

(27) Fremond, M. L. et al. Overview of STING-Associated Vasculopathy with Onset in Infancy (SAVI) Among 21 Patients. J Allergy Clin Immunol Pract 9, 803-818 e811 (2021).

Version 2

Abstract

Abstract

The significance of STING (encoded by the TMEM173 gene) in tissue inflammation and cancer immunotherapy has been increasingly recognized. Intriguingly, common human TMEM173 alleles R71H-G230A-R293Q (HAQ) and G230A-R293Q (AQ) are carried by ∼60% of East Asians and ∼40% of Africans, respectively. Here, we examine the modulatory effects of HAQ, AQ alleles on STING-associated vasculopathy with onset in infancy (SAVI), an autosomal dominant, fatal inflammatory disease caused by gain-of-function human STING mutations. CD4 T cellpenia is evident in SAVI patients and mouse models. Using STING knock-in mice expressing common human TMEM173 alleles HAQ, AQ, and Q293, we found that HAQ, AQ, and Q293 splenocytes resist STING-mediated cell death ex vivo, establishing a critical role of STING residue 293 in cell death. The HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice did not have CD4 T cellpenia. The HAQ/SAVI(N153S), AQ/SAVI(N153S) mice have more (∼10-fold, ∼20-fold, respectively) T-regs than WT/SAVI(N153S) mice. Remarkably, while they have comparable TBK1, IRF3, and NFκB activation as the WT/SAVI, the AQ/SAVI mice have no tissue inflammation, regular body weight, and normal lifespan. We propose that STING activation promotes tissue inflammation by depleting T-regs cells in vivo. Billions of modern humans have the dominant HAQ, AQ alleles. STING research and STING-targeting immunotherapy should consider TMEM173 heterogeneity in humans.

Introduction

STING drives cytosolic DNA-induced type I IFNs production 1. Recent research revealed that STING promotes inflammation in a variety of inflammatory diseases, including nonalcoholic fatty liver disease, nonalcoholic steatohepatitis, kidney injury, neurodegenerative diseases, cardiovascular diseases, obesity, diabetes, and aging 2–9. The type I IFNs-independent function of STING has also emerged 9–12. For example, initially described as a type I interferonopathy 13, recent studies in STING-associated vasculopathy with onset in infancy (SAVI) mouse models showed that SAVI is largely independent of type I IFNs 14–17. In a STING N153S mouse model of SAVI, crossing N153S mice to IRF3/IRF7, and IFNAR1 knockout mice, N153S mice still developed spontaneous lung diseases 14. JAK inhibitors were used to block type I IFNs signaling for SAVI patients with mixed success 17–20. For example, in a review of JAK inhibition in 18 SAVI patients, incomplete response to treatment happened in 7/18 (38%) of patients 21. Furthermore, 2 patients died of respiratory failure despite this treatment 21. Both radioresistant parenchymal and/or stromal cells and hematopoietic cells influence SAVI pathology in mice 22,23. The observation is important because it predicts that allogeneic stem cell transplantation may not work in human SAVI patients. Indeed, lung transplantations did not show improvement in SAVI patients 18,24. Patients died at 3- and 9 months post-lung transplant 18,24. How STING drives inflammation in vivo, independent of type I IFNs, remains unknown. Consequently, there is no curative care for SAVI.

Characterized as an innate immune sensor, STING expression is, paradoxically, high in CD4 T cells 13,25. Furthermore, STING activation kills mouse and human CD4 T cells ex vivo 26–28. SAVI patients and mouse models had CD4 T cellpenia 13,28. STING was first discovered as MPYS for its cell growth inhibition and cell death function in mouse B lymphoma cells 29. STING-mediated cell death is cell type dependent. For example, while STING activation kills human endothelial cells, primary and cancerous T cells, it does not kill mouse MEFs, BMDCs, or BMDMs 30–32. Second, STING-mediated cell death is type I IFNs-independent 28,30,32. Multiple cell death pathways, i.e., apoptosis, necroptosis, pyroptosis, ferroptosis, and PANoptosis, are proposed 32–34. Last, the in vivo biological significance of STING-mediated CD4 T cell death is not clear 28,32. In humans, SAVI patients with constitutively activated STING have low CD4 T cell numbers 13, and type I IFNs are dispensable for STING-mediated human CD4 T cell death 28. Different from SAVI mice, SAVI patients (N154S or V155M) had normal counts of CD8 T and B cells 13.

The human TMEM173 gene is highly heterogeneous 35,36. ∼50% of people in the U.S. carry at least one copy of non-WT TMEM173 allele 35. Among them, the R71H-G230A-R293Q (HAQ) is the 2nd most common TMEM173 allele carried by ∼23% of people in the U.S. 35. However, in East Asians, WT/HAQ (34.3%), not WT/WT (22.0%), is the most common TMEM173 genotype 36. Critically, the HAQ allele was positively selected in modern humans outside Africa 37. Anatomically modern humans outside Africa are descendants of a single Out-of-Africa Migration 50,000∼70,000 years ago. ∼1.4% of Africans have the HAQ allele, while ∼63.9% of East Asians are HAQ carriers 37. Haplotype analysis revealed that HAQ was derived from G230A-R293Q (AQ) allele 37. Importantly, the AQ allele was negatively selected outside Africa 37. ∼40.1% of Africans are AQ carriers, while ∼0.4% of East Asians have the AQ allele 37. TMEM173 alleles often have a dominant negative effect likely because the protein STING exists as a homodimer 29,38. SAVI is an autosomal dominant inflammatory disease 13. WT/HAQ individuals had reduced Pneumovax23®-induced antibody responses compared to WT/WT individuals (NCT02471014) 39. Notably, AQ responds to CDNs and produces type I IFNs in vivo and in vitro 37,40,41, but the AQ allele was negatively selected in non-Africans 37. In contrast, the HAQ allele, defective in CDNs-type I IFNs responses 35,36,39,42,43, was positively selected in non-Africans 37, indicating that the CDNs-type I IFNs independent function of STING was essential for the survival of early modern humans outside of Africa.

In this study, we discovered, surprisingly, that the HAQ, AQ splenocytes are resistant to STING-mediated cell death. We generated HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice and found that the HAQ, AQ alleles prevent CD4 T cellpenia, increasing/restoring T-regs and alleviating/stopping tissue inflammation in SAVI mice, thus providing evidence for the in vivo significance of type I IFNs-independent, STING-mediated cell death and potential AQ-based curative therapy for SAVI patients.

Results

STING activation kills mouse spleen CD4, CD8 T, and CD19 B cells ex vivo

We first used the synthetic non-CDNs STING agonist diABZI 44 to induce lymphocyte death because diABZI induces cell death without the need for lipid transfection or detergent for cell permeabilization 31,45 and diABZI is in clinical trials (NCT05514717). Splenocytes from C57BL/6N mice were treated with diABZI in culture, and cell death was determined by Annexin V and Propidium Iodide stain. Splenocyte cell death could be detected as early as 5 hours post diABZI treatment (Figure S1A). Dosage responses showed that ∼25ng/ml diABZI could kill 70% of splenocytes (Figure S1B). Similarly, STING agonists DMXAA and synthetic CDNs RpRpss-Cyclic di-AMP killed mouse spleen CD4, CD8 T cells, and CD19 B cells (Figure 1A). Thus, STING activation readily induces mouse lymphocyte death ex vivo.

Splenocytes from HAQ, AQ, and Q293 mice are resistant to STING-mediated cell death ex vivo.

(A). C57BL/6N splenocytes were treated directly (no transfection) with diABZI (100ng/ml), RpRpss-Cyclic di-AMP (5 μg/ml) or 2′3′-cGAMP (10μg/ml), DMXAA (25μg/ml) for 24hrs in culture. CD4, CD8 T cells and CD19 B cells death were determined by PI staining. (B). Splenocytes from C57BL/6N mice were pre-treated with indicated small molecules, GSK2334470 (1.25µM), GSK8612 (2.5µM), Bx-795 (0.5µM), QVD-OPH (25µM) for 2hrs. diABZI (100ng/ml) was added in culture for another 24hrs. Dead cells were determined by PI staining. (C-D). Flowcytometry of HAQ, AQ, IFNAR1-/- or C57BL/6N splenocytes treated with diABZI (100ng/ml) for 24hrs. Cell death was determined by PI staining. (E-F). Q293 or the WT littermates splenocytes were treated with diABZI (100ng/ml), RpRpss-Cyclic di-AMP (5 μg/ml) or 2′3′-cGAMP 10μg/ml) for 24h. Cell death was determined by PI staining. (G). WT/HAQ, WT/AQ, or WT/WT littermates splenocytes were treated with DMXAA (10, 25 or 100µg/ml) for 24h. Cell death was determined by PI staining. Data are representative of three independent experiments. Graphs represent the mean with error bars indication s.e.m. p values are determined by one-way ANOVA Tukey’s multiple comparison test (A, E, G) or unpaired student T-test (B, D). * p<0.05. n.s: not significant.

TBK1 activation is required for STING-mediated mouse spleen cell death ex vivo

STING activation can lead to apoptosis, pyroptosis, necroptosis, or ferroptosis 28–34,45,46. We then treated mouse splenocytes with apoptosis, pyroptosis, necroptosis inhibitors, STING inhibitors H-151, C-176, and palmitoylation inhibitor 2-bromopalmitate (2-BP), followed by diABZI stimulation. Inhibitors for NLRP3 (MCC950), RIPK1 (Necrostatin-1), RIPK3 (GSK872), Caspase-1 (VX-795), Caspase-3 (Z-DEVD-FMK), Caspase 1,3,8,9 (Q-VD-Oph), ferroptosis (liproxstatin-1) did not affect diABZI-induced splenocyte cell death ex vivo (Figure S1B, S1C). The STING inhibitors H-151, C-176, and 2-BP also could not prevent diABZI-induced cell death (Figure S1C) though they inhibited diABZI-induced IFNβ production (Figure S1D). Instead, the TBK1 inhibitor BX-795 abolished diABZI-induced splenocyte death (Figure S1C).

BX-795 is a multi-kinase inhibitor, including 3-phosphoinositide-dependent protein kinase 1 (PDK1) and TBK1 (IC50s = 6 and 11 nM, respectively). However, the treatment of PDK1 inhibitor GSK2334470 (IC50 = 10nM) did not prevent diABZI-induced splenocyte death (Figure 1B). In contrast, GSK8612, a highly potent and selective inhibitor for TBK1, prevented diABZI-induced splenocyte death (Figure 1B). Thus, TBK1 activation is likely critical for STING-mediated splenocyte cell death ex vivo.

HAQ, AQ, Q293 STING knock-in mouse splenocytes are resistant to STING-mediated cell death ex vivo

HAQ and AQ are common human TMEM173 alleles 35–37. Previously, we reported that HAQ knock-in mice are defective in CDNs-induced immune responses, while CDNs responses in AQ knock-in mice are similar to WT mice 37. We treated splenocytes from HAQ and AQ mice with diABZI ex vivo and found, surprisingly, that both HAQ and AQ splenocytes were resistant to diABZI-induced cell death (Figure 1C, 1D). In comparison, IFNAR1-/- splenocytes were killed by diABZI, confirming that STING-mediated lymphocytes death are type I IFNs-independent (Figure 1C, 1D) 28,30,32.

HAQ and AQ share the common A230 and Q293 residues changes. We thus generated a Q293 TMEM173 knock-in mouse. Notably, the Q293 splenocytes were resistant to STING agonists 2’3’-cGAMP, RpRpss-Cyclic di-AMP, and diABZI-induced cell death (Figure 1E, 1F). Thus, the residue 293 of STING is critical for its cell death function.

WT/HAQ, WT/AQ mouse splenocytes are partially resistant to STING-mediated cell death ex vivo

WT/HAQ (34.3%) is the most common human TMEM173 genotype in East Asians, while WT/AQ (28.2%) is the 2nd most common TMEM173 genotype in Africans 36. We generated WT/HAQ, WT/AQ mice and treated their splenocytes with mouse STING agonist DMXAA. WT/HAQ and WT/AQ splenocytes were protected from 25µg/ml DMXAA-induced cell death (Figure 1G). 100µg/ml DMXAA could kill WT/HAQ and WT/AQ splenocytes, albeit less than WT/WT cells (Figure 1G). Thus, the HAQ and AQ alleles are dominant and likely impact STING activation even in heterozygosity.

STING activation kills primary human CD4 T cells but not CD8 T or CD19 B cells

STING agonists-based clinical trials in humans have been disappointing (NCT02675439, NCT03010176, NCT05514717) 47,48. We showed that the human TMEM173 gene might undergo natural selection during the out-of-Africa migration 37 sensitive to evolutionary pressure. Thus, we investigated STING-mediated death in primary human lymphocytes.

Human explant lung cells from the WT(R232)/WT(R232) donors were treated with STING agonists 2’3-c GAMP, RpRpss-Cyclic di-AMP, diABZI for 24 hours in culture. Lymphocyte cell death was determined by Propidium Iodide staining. Different from mouse lymphocytes, diABZI and RpRpss-Cyclic di-AMP killed human CD4 T but not CD8 T or CD19 B cells (Figure 2A, 2B). Human CD8 T and CD19 B cells are resistant to 500ng/ml diABZI-induced cell death (Figure S2A).

HAQ, AQ, and Q293 human cells are resistant to STING agonists-induced death.

(A-B). Total human Lung cells from WT/WT individuals were treated with diABZi (100ng/ml) for 24hrs. Cell death in CD4, CD8 T cells and CD19 B cells were determined by PI staining. (C). Total lung cells from a WT/HAQ (2 individuals) and a WT/WT (3 individuals) were treated with diABZi (25, 100ng/ml) for 24h. Cell death in CD4 T cells was determined by PI staining. (D-E). STING-KO THP-1 cells (Invivogen,, cat no. thpd-kostg) were stably reconstituted with human WT (R232), HAQ, AQ, Q293. Cells were treated with diABZI (200ng/ml) in culture for 24 hrs. Dead cells were determined by Annexin V staining. (F). STING-KO THP-1 cells stably reconstituted with human WT (R232), Q293 were treated with indicated dose of diABZI for 24hs in culture. Dead cells were determined by Annexin V staining. (G). STING-KO THP-1 cells stably reconstituted with human WT (R232), Q293 were treated with indicated dose of diABZI for 2hs in culture. STING activation was detected by anti-STING antibody (Proteintech, #19851-1-AP). (H). STING-KO THP-1 cells stably reconstituted with human WT (R232), HAQ, AQ, Q293 were treated with 50ng/ml diABZI in culture for 24hrs. ISG-54 reporter luciferase activity was determined in cell supernatant and normalized to 10ng/ml IFNβ-stimulated ISG-54 luciferase activity. (I). STING-KO THP-1 cells stably reconstituted with human WT (R232), HAQ, AQ were treated with 50ng/ml diABZI in culture for 2hrs. STING and IRF3 activation were determined by anti-STING antibody and anti-p IRF3 antibody (CST, Ser396, clone 4D4G). Densitometry was determined by ImageLab 5. Data are representative of three independent experiments. Graphs represent the mean with error bars indication s.e.m. p values determined by one-way ANOVA Tukey’s multiple comparison test (B, C, F, H, G) or unpaired student T-test (E, I). * p<0.05, n.s: not significant.

WT/HAQ human CD4 T cells are resistant to low doses of diABZI-induced cell death

WT/HAQ mouse splenocytes are resistant to low-dose diABZI-induced cell death (Figure 1G). To extend our observation into primary human T cells, we obtained lung explants from WT/WT and WT/HAQ individuals (Figure S2B) and treated them with diABZI in culture. 25ng/ml diABZI killed WT/WT, but not WT/HAQ, human lung CD4 T cells (Figure 2C).

diABZI induces cell death in STING-KO human THP-1 cells reconstituted with WT human STING (R232) but not HAQ, AQ or Q293 human STING allele

To further determine cell death influenced by human TMEM173 alleles HAQ, AQ, and Q293, we used the STING-KO THP-1 cell line because STING agonist induces type I IFNs and cell death in STING-KO THP-1 cells expressing WT human STING 34 (Figure S2C, S2D). We, thus, generated stable THP-1 STING-KO lines expressing HAQ, AQ, WT, or Q293 TMEM173 allele. Cell death was determined by Annexin-V staining. diABZI killed THP-1 STING-KO lines expressing WT but not HAQ, AQ, or Q293 TMEM173 allele (Figure 2D, 2E). No cell death was induced in the Q293 THP-1 cells stimulated by 20 to 200 ng/ml of diABZI (Figure 2F). diABZI also did not induce STING activation in Q293 THP-1 cells (Figure 2G). Notably, 50ng/ml diABZI induced p-IRF3 activation and type I IFNs in AQ THP-1 cells but not HAQ THP-1 cells (Figure 2H, 2I), indicating that the STING-cell death and STING-IRF3-Type I IFNs pathways can be uncoupled.

HAQ and AQ alleles rescue the lymphopenia and suppress myeloid cell expansion in SAVI(N153S) mice.

The in vivo significance of the STING/MPYS-cell death is unclear. Furthermore, multiple cell death pathways, i.e., apoptosis, necroptosis, pyroptosis, ferroptosis, and PANoptosis, are proposed 32–34. The uncertainty likely results from studies using different cell types (primary cells vs cancer cell lines); species (human vs mouse); STING agonists (cGAMP, which requires cell permeabilization by detergents or lipid transfection, vs diABZi, DMXAA that can directly cross the membrane) 26,27,30,31,49. Critically, which mechanism is relevant in vivo, causing T cellpenia is not known. To clarify the in vivo significance and mechanisms of STING-mediated cell death, we turned to SAVI mice.

SAVI is an autosomal dominant, inflammatory disease caused by one copy of a gain-of-function TMEM173 mutant (WT/SAVI) 13. CD4 T cellpenia was found in SAVI patients and SAVI mouse models 13,14. STING activation in SAVI mice is independent of ligands and happens in vivo. We thus generated HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice aiming to establish the in vivo significance and mechanism of STING-cell death.

First, HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice had reduced splenomegaly compared to WT/SAVI(N153S) mice though their spleens were still larger than the littermates WT/HAQ and WT/AQ (Figure 3A, 3B). Next, HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice had similar spleen B cells and CD4 T cell numbers as the WT/HAQ, WT/AQ littermates (Figure 2B, 2C). Their CD8+ T cells were lower than their WT littermates but much higher than the WT/SAVI(N153S) mice (Figure 3D). Third, spleen myeloid cell numbers, i.e., neutrophils, Ly6Chi monocytes and F4/80 macrophages, were all reduced by half compared to WT/SAVI(N153S) mice (Figure 3E∼3H). Last, splenocytes from HAQ/SAVI, AQ/SAVI mice were partially resistant to diABZI, DMXAA-induced cell death ex vivo (Figure S3). Notably, the HAQ/SAVI(N153S) and AQ/SAVI(N153S) mice also had restored bone marrow monocytes (Figure S4). Thus, HAQ and AQ alleles prevent lymphopenia and suppress myeloid cell expansion in SAVI(N153S) mice.

HAQ and AQ rescue the lymphopenia and suppress myeloid cell expansion in SAVI (N153S) mice.

(A). The size and weight of spleens from WT/HAQ, HAQ/SAVI, AQ/SAVI, WT/AQ, WT/SAVI. (B-D). Spleen CD19+ B cells, CD4, CD8 T cells were determined in the indicated mice by Flow. (E-H). Spleen Ly6G+ neutrophils, Ly6Chi monocytes and F4/80+ macrophage was determined in the indicated mice by Flow. Data are representative of three independent experiments. n=3∼5 mice/group. Graphs represent the mean with error bars indication s.e.m. P values are determined by one-way ANOVA Tukey’s multiple comparison test. * p<0.05, n.s: not significant.

The HAQ allele alleviates and the AQ allele prevents SAVI(N153S) disease in mice

SAVI(N153S) disease is characterized by early onset, failure to thrive (low body weight), persistent lung inflammation, decreased lung function, and young death in humans and mouse models 13,17,49,50. The HAQ/SAVI(N153S) mice weighed more and had an improved lifespan than the WT/SAVI(N153S) mice (Figure 4A, 4B). The lifespan, airway resistance, and tissue inflammation (lung, liver) were also improved in HAQ/SAVI(N153S) mice compared to the WT/SAVI(N153S) mice (Figure 4C, 4D, 4J, 4K). However, the pulmonary artery pressure was still elevated in HAQ/SAVI(N153S) mice (Figure 4E). Remarkably, the AQ/SAVI(N153S) mice had similar body weight and lifespan as the WT/AQ mice (Figure 4F, 4G). The airway resistance, pulmonary artery pressure, and tissue inflammation in AQ/SAVI(N153S) were similar to the WT/AQ littermates (Figure 4H, 4I, 4J, 4K). Thus, the HAQ allele alleviates and the AQ allele prevents inflammatory SAVI disease in mice. Interestingly, lungs from HAQ/SAVI, AQ/SAVI had similarly reduced infiltration of Ly6G+CD11B+ neutrophils and Ly6G-Ly6C+CD11B+ inflammatory monocytes compared to WT/SAVI mice (Figure S5).

HAQ and AQ alleles prevent SAVI(N153S) disease in mice.

(A, B, G). The size and body weight of HAQ/SAVI, AQ/SAVI, WT/SAVI and their littermates WT/HAQ, WT/AQ mice. (D, E, H, I). Airway resistance, and pulmonary artery pressure were determined as described in Materials and Methods. (C, F). HAQ/SAVI, AQ/SAVI, WT/SAVI (10 mice/group), were monitored for survival by Kaplan-Meier. (J, K). Representative hematoxylin and eosin (H&E) staining of lung, liver sections from indicated mice. n=3∼5 mice/group. Data are representative of 3 independent experiments. Graphs represent the mean with error bars indication s.e.m. p values are determined by one-way ANOVA Tukey’s multiple comparison test. * p<0.05, **p<0.01. n.s.: not significant. WBC: white blood cells; H: hepatocytes; K: Kupper cells; PV: portal vein; CV: central vein.

diABZI induces similar STING, TBK1, IRF3, NFκB activation in the AQ/SAVI(N153S) and WT/SAVI(N153S) bone-marrow-derived-macrophage (BMDM)

SAVI was characterized as type I interferonopathy 17. However, several studies showed that type I IFN signaling and IRF3 activation were dispensable for SAVI disease 14,15,23,50. AQ allele prevents SAVI disease (Figure 4). However, diABZI-treated AQ/SAVI(N153S) and WT/SAVI(N153S) BMDM had similar TBK1-IRF3 activation and IFNβ production (Figure 5A, 5B). diABZI treatment caused IκBα degradation, and similar TNF production in WT/SAVI and AQ/SAVI BMDM (Figure 5C, 5D). Furthermore, diABZI activation led to STING protein degradation in WT/SAVI and AQ/SAVI BMDM (Figure 5E). Last, using cleavable crosslinker dithiobis (succinimidyl propionate (DSP), we showed that STING in WT/SAVI, AQ/SAVI BMDM forms a similar dimer in situ (Figure 5F). Thus, the AQ/SAVI BMDM had similar STING degradation, TBK1, IRF3, NFκB activation, and dimerization as the WT/SAVI BMDM.

AQ/SAVI(N153S) cells had similar TBK1-IRF3, NFκB activation and STING degradation as the WT/SAVI(N153S) cells.

(A, C). BMDM from WT/WT, WT/SAVI, HAQ/SAVI and AQ/SAVI were treated with 100ng/ml diABZi in culture for 2 hrs. Cells were lysed and run on a 4∼20% Mini-PROTEAN TGX Stain-Free Precast Gel. The blot was probed for phospho-Thr172-TBK1 antibody (CST, clone D52C2), phosphor-Ser 396-IRF3 (CST, clone 4D4G), STING (Proteintech, #19851-1-AP) and IκBα (CST, clone 44D4) antibody. (E). BMDM from WT/WT, WT/SAVI, HAQ/SAVI and AQ/SAVI were treated with 100ng/ml diABZi in culture for 24 hrs. Cells were lysed and run on a 4∼20% Mini-PROTEAN TGX Stain-Free Precast Gel. The blot was probed for STING antibody (Proteintech, 19851-1-AP). (B, D). IFNβ and TNF were determined by ELISA in the cell supernatant from E. (F). BMDM from WT/WT, WT/SAVI, HAQ/SAVI and AQ/SAVI were treated with 400µM cleavable chemical crosslinker DSP (Pierce™, cat no: PG82081) in PBS for 1 hr at 4°C. Cells were washed with PBS and lysed in RIPA buffer. Whole cell lysate was mixed with 4x Laemmli Sample Buffer (BioRad, cat no 1610747) containing 5% 2-mercaptoethanol, heated at 95°C for 10 min and, run on a 4∼20% Mini-PROTEAN TGX Stain-Free Precast Gel. The blot was probed for STING antibody (Proteintech, 19851-1-AP). Densitometry was determined by ImageLab 5. Data are representative of three independent experiments. Graphs represent the mean with error bars indication s.e.m. p values are determined by unpaired student T-test (A-E) or one-way ANOVA Tukey’s multiple comparison test (D, B). * p<0.05, **p<0.01, ***p<0.001. n.s.: not significant; n.d: not detected.

The HAQ allele increased, and the AQ allele restored T-regs in SAVI(N153S) mice

IFNγ was proposed to drive SAVI disease 15,23,38. We confirmed that WT/SAVI CD4 T cells were enriched with IFNγ+ cells (Figure 6A). However, WT/SAVI mice have CD4 T cellpenia. Thus, the total numbers of spleen IFNγ+ CD4 T cells were comparable in WT/SAVI and AQ/SAVI mice (Figure 6A). In contrast, the HAQ/SAVI mice had decreased IFNγ+ CD4 T cells (Figure 6A).

HAQ/SAVI(N153S) and AQ/SAVI(N153S) cells had 10-fold and 20-fold increased spleen T-regs compared to WT/SAVI mice.

(A). Flow cytometry analysis of IFNγ producing spleen CD4+ T cells from WT/HAQ, WT/AQ, WT/SAVI, HAQ/SAVI and AQ/SAVI mice. (B). Flow cytometry analysis of CD4+ FoxP3+ spleen T-regs. n=3∼5 mice/group. Data are representative of 3 independent experiments. Graphs represent the mean with error bars indication s.e.m. p values are determined by one-way ANOVA Tukey’s multiple comparison test. * p<0.05, **p<0.01, ***p<0.001. n.s.: not significant.

The induction of Foxp3 expression in T-reg cells during ongoing autoimmune inflammation resolved inflammation and pathology in mice 51. CD4 T cellpenia depletes CD4 T-regs. Indeed, WT/SAVI mice had ∼20-fold reduction of spleen FoxP3+ T-regs compared to AQ/SAVI or WT/WT littermates (Figure 6B). The HAQ/SAVI mice also had ∼10-fold more T-regs than the WT/SAVI littermate (Figure 6B).

Discussion

This study, using the HAQ, AQ, SAVI(N153S) TMEM173 knock-in mice, reveals the in vivo significance and mechanism of STING-mediated CD4 T cell death. HAQ, AQ alleles prevent CD4 T cellpenia, and increase/restore CD4 T-regs in SAVI mice. The results are consistent with previous finding that the impaired CD4 T cell proliferation by the SAVI(V155M) mutant could be rescued by the addition of the HAQ allele in vitro 27. STING has been increasingly implicated in inflammatory diseases such as nonalcoholic fatty liver disease, nonalcoholic steatohepatitis, cardiomyopathy, obesity, diabetes, neurodegenerative diseases, aging, and kidney injury, many of which are independent of type I IFNs 2,3,9. It is tempting to suggest that STING activation in CD4 T cells leads to CD4 T-regs depletion that break tissue tolerance and exacerbates tissue inflammation.

Human immunodeficiency virus (HIV) primarily infects CD4 T cells and might activate the STING pathway in CD4 T cells 52–57. The loss of CD4 T cells is the hallmark of untreated HIV infection 58,59, and the measurement of CD4 T cell count is a central part of HIV care. We found that HAQ and AQ CD4 T cells are resistant to STING-mediated cell death. Mogensen and colleagues reported that HAQ/HAQ was enriched in HIV-infected long-term nonprogressors 42. These HAQ/HAQ individuals had reduced inhibition of CD4 T cell proliferation and a reduced immune response to DNA and HIV 42. It is likely that HIV infection activates STING-cell death pathway in CD4 T cells. In HAQ/SAVI and AQ/SAVI mice, one copy of HAQ, AQ allele suppressed CD4 T cell death. HAQ, AQ carriers might have fewer HIV-induced CD4 T cell death, thus being long-term nonprogressors in HIV infection-induced Acquired immunodeficiency syndrome (AIDS)42. Targeting STING to prevent CD4 T cell death might be a valid therapy for AIDS. Activating the STING pathway is a promising strategy for cancer immunotherapy 60–66.

Multiple STING agonists are in clinical trials 47,48. Recently, the safety issue emerged in some STING agonist trials 47,48. For example, in the STING agonist, ADU-S100, clinical trial, Grade 3/4 treatment-related adverse events were reported in 12.2% of 41 pretreated patients (NCT02675439)47,48. The National Institutes of Health defines grade 3 as “incapacitating; unable to perform usual activities; requires absenteeism or bed rest.” In a clinical trial using STING antibody-drug-conjugate (ADC) that conjugates diABZI to anti-HER2 Ab, a Grade 5 (fatal) serious adverse event was recorded and deemed related to the STING-ADC (NCT05514717). SAVI disease, driven by overreacting STING, is often fatal13. AQ, to a less degree, HAQ, suppress mortality in SAVI mice. Future STING clinical trials should be based on human TMEM173 genotype to achieve safe and effective responses.

Mechanistically, apoptosis, pyroptosis, ferroptosis, necroptosis, and PANoptosis have all been reported in STING-mediated cell death 28–34,45,46. Different cell types and STING agonists used likely contributed to the inconsistency and complexity. Here, we focused on lymphopenia in the SAVI mice that avoids ligand-dependent, non-physiological dosage in STING-mediated cell death. HAQ and AQ alleles could prevent CD4 T cellpenia in the SAVI mice strongly indicating that residue A230 or Q293 prevent STING-mediated CD4 T cell death in vivo. Splenocyte from Q293 mice were resistant to STING agonists-induced cell death ex vivo. Thus, it is likely that the Q293 residue is critical for STING-mediate lymphopenia. Notably, Q293 is outside the C-terminal tail (CTT) (residues 341–379 of human STING) critical for TBK1 recruitment and IRF3 phosphorylation 67 or miniCTT domain (aa343–354) 27, or the UPR motif (aa322–343) 49 important for T cells death in vitro. Further studies are needed to understand how the aa293 of STING mediates cell death in vivo. Noteworthy, AQ/SAVI cells had similar TBK1-IRF3, NFκB activation and STING degradation as the WT/SAVI cells. Yet, AQ/SAVI mice did not have CD4 T cellpenia as WT/SAVI mice suggesting that the canonical STING-TBK1-IRF3/NFκB pathway, likely STING oligomerization, is not sufficient for the induction of cell death at the physiological condition.

We used the WT/N153S knock-in SAVI mouse model that spontaneously develop lung inflammation, T cell cytopenia, and early mortality, mimicking pathological findings in human SAVI patients16. Using the WT/N153S SAVI mouse model and human Jurkat T cell line, it was proposed that STING activation causes chronic ER stress and unfolded protein response, leading to T cell death by apoptosis 49. Furthermore, the study showed that crossing WT/N153S mice to the OT-I mice reduced ER stress and restored CD8+, but not CD4+, T cells 49. The restoration of CD8+T cells reduces inflammation and lung disease 49. However, human WT/N154S SAVI patients have normal CD8+ T cells numbers 13, and primary human CD8+ T cells are largely resistant to STING-agonists-induced cell death ex vivo (Figure 2A) 28. Thus, it is puzzling how restoring CD8+ T cells can rescue SAVI phenotypes since the SAVI patients already have normal CD8+ T cells numbers.

Finally, it is unexpected that both HAQ and AQ alleles are resistant to cell death. Our previous studies showed that the HAQ and AQ alleles have opposite functions 37. AQ-STING, not HAQ-STING, responds to CDNs 35–37,39–43,68. AQ mice are lean while HAQ mice are fat 37. Most importantly, HAQ was positively selected, while AQ was negatively selected, in modern humans outside Africans 37. Thus, the death pathway of STING is also distinct from the STING function that was naturally selected. It is worth noting that the AQ allele does better than the HAQ allele in suppressing SAVI disease. Thus, besides preventing cell death, additional mechanism by the AQ allele, e.g. fatty acid metabolism 37,69, is involved in curing SAVI.

The limitations of the study

The poor transferability of mouse to humans is a major issue in STING research 47,48. The present study used AQ/SAVI and HAQ/SAVI mice. Confirmation is needed in humans with the identification and evaluation of people who are AQ/SAVI, HAQ/SAVI.

Materials and Methods

Experimental Design

The study was designed to reveal (i) the in vivo significance of the type I IFNs-independent, STING-dependent cell death function; (ii) the interplay between common STING alleles HAQ, AQ and the rare, gain-of-function SAVI STING mutation; (iii) the driver for the inflammatory SAVI disease. Mouse splenocytes, primary human lung cells, human THP-1 cells and HAQ, AQ, SAVI knock-in mice were used to establish the in vivo significance and human relevance. All the repeats were biological replications that involve the same experimental procedures on different mice. Where possible, treatments were assigned blindly to the experimenter by another individual in the lab. When comparing samples from different groups, samples from each group were analyzed in concert, thereby preventing any biases that might arise from analyzing individual treatments on different days. All experiments were repeated at least twice.

Mice

WT/SAVI(N153S) mice were purchased from The Jackson Laboratory. HAQ, AQ mice were previously generated in the lab 36,37. The Q293 mice were generated by Cyagen Biosciences. Briefly, the linearized targeting vector was transfected into JM8A3.N1 C57BL/6N embryonic stem cells. A positive embryonic stem clone was subjected to the generation of chimera mice by injection using C57BL/6J blastocysts as the host. Successful germline transmission was confirmed by PCR sequencing. The heterozygous mice were bred to Actin-flpase mice [The Jackson Laboratory, B6.Cg-Tg (ACTFLPe)9205Dym/J] to remove the neo gene and make the Q293 knock-in mouse. Age-and gender-matched mice (2 – 6 month old, both male and female) were used for indicated experiments. WT/SAVI (male) x WT/HAQ (female), WT/SAVI (male) x WT/AQ (female) breeders were set up to generate HAQ/SAVI, AQ/SAVI mice. Mice were housed at 22°C under a 12-h light-dark cycle with ad libitum access to water and a chow diet (3.1 kcal/g, Teklad 2018, Envigo, Sommerset, NJ) and bred under pathogen-free conditions in the Animal Research Facility at the University of Florida. Littermates of the same sex were randomly assigned to experimental groups. All mouse experiments were performed by the regulations and approval of the Institutional Animal Care and Use Committee at the University of Florida, IACUC202200000058.

Reagents

Recombinant human IFNβ (R&D, cat no. 8499-IF-010/CF), diABZI (Invivogen, cat no. 2138299-34-8), 2’3’-cGAMP (Invivogen, cat no. tlrl-nacga23-02), DMXAA (Invivogen, cat no. tlrl-dmx), H151 (Invivogen, cat no. inh-h151), RpRpSS-Cyclic di-AMP (Biolog, cat no. C118), THP1-Dual™ KO-STING Cells (Invivogen, cat no. thpd-kostg). All other chemical inhibitors are from Selleckchem. Mouse TNF alpha ELISA Ready Set Go. (eBioscience, cat no. 88-7324). Mouse IFN-Beta ELISA Kit (PBI, cat no. 42400).

Generation of THP-1 KO-STING cells stably expressing human TMEM173 alleles

THP1-Dual™ KO-STING Cells in 6-well plate were transfected with 1µg TMEM173 plasmid (in pcDNA 3.1 vector) with Lipofectamine LTX and Plus Reagent (Invitrogen, cat no: A12621) according to the manufacturer’s instructions. Transfecting Plasmid DNA into THP-1 Cells Using Lipofectamine LTX Reagent | Thermo Fisher Scientific - US. 48hrs after the transfection, the cell medium was changed. G418 (1mg/ml) was added to the culture to select STING expressing THP-1 cells. The G418-resistant cells were established and expanded.

Histology

Lungs and livers were fixed in 10% formalin, paraffin-embedded, and cut into 4-µm sections. Lung, liver sections were then stained for hematoxylin-eosin. All staining procedures were performed by the histology core at the University of Florida. Briefly, tissue sectins were immersed Harris Hematoxylin for 10 seconds, then washed with tap water. Cleard sections were re-immersed in EOSIN stain for ∼30 seconds. The sections were washed with tap water until clear, then dehydrate in ascending alcohol solutions (50%,70%,80%,95% x 2, 100% x 2). Afterwards, the sections werer cleared with xylene (3 - 4 x). The sections were mounted on glass slide with permount organic mounting medium for visulization.

Lung Function

Pulmonary function was evaluated using an isolated, buffer-perfused mouse lung apparatus (Hugo Sachs Elektronik, March-Huggstetten, Germany), as previously described 70. Briefly, mice were anesthetized with ketamine and xylazine and a tracheostomy was performed, and animals were ventilated with room air at 100 breaths/min at a tidal volume of 7 μl/g body weight with a positive end-expiratory pressure of 2 cm H2O using a pressure-controlled ventilator (Hugo Sachs Elektronik, March-Huggstetten, Germany).

Isolation of lung cells

Cells were isolated from the lung as previously described 71. The lungs were perfused with ice-cold PBS and removed. Lungs were digested in DMEM containing 200μg/ml DNase I (Roche, 10104159001), 25μg/ml Liberase TM (Roche, 05401119001) at 37°C for 2 hours. Red blood cells were then lysed and a single cell suspension was prepared by filtering through a 70-µm cell strainer.

BMDM activation

BMDMs were induced from mouse bone marrow cells cultured in RPMI 1640 (cat#11965; Invitrogen) with 10% FBS, 2 mM L-glutamine, 1 mM sodium pyruvate, 10 mM HEPES buffer, 1% nonessential amino acids, 50 mM 2-ME, 1% Pen/Strep, with 20ng/ml M-CSF (Kingfisher, RP0407M) for 10 days 36. STING agonists were added into the culture (no transfection or membrane permeabilization).

Flow cytometry

Single cell suspensions were stained with fluorescent-dye-conjugated antibodies in PBS containing 2% FBS and 1mM EDTA. Surface stains were performed at 4°C for 20 min. For intracellular cytokine or transcription factor staining of murine and human cells, cells were fixed and permeabilized with the Foxp3 staining buffer set (eBioscience, cat no 00-5523-00). Cells were washed and stained with surface markers. Cells were then fixed and permeabilized (eBioscience, cat no. 00-5523-00) for intracellular cytokine stain. Data were acquired on a BD LSRFortessa and analyzed using the FlowJo software package (FlowJo, LLC). Cell sorting was performed on the BD FACSAriaIII Flow Cytometer and Cell Sorter. More information on the antibodies used can be found in supplementary table S1.

Human lung explants

Human lung explants were procured at the Lung Transplant Center, Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University of Florida. Donor and patients consent was obtained for a research protocol (UF IRB201902955-Treatment with IFNβ Induces Tolerogenic Lung Dendritic Cells in Human advanced lung disease). Healthy donor lungs were surgically removed postmortem, perfused, small pieces were cut from the right middle and lower lobes for research purpose, and stored in cold Perfadex ® at 4°C for no more than 12 hrs before processing. Ex planted lungs from emphysema lung transplant patients were stored in cold Perfadex ® at 4°C for no more than 12 hrs before the process. No lung explants were procured from prisoners.

Statistical Analysis

To gain statistical power, we employ three∼four mice/groups to characterize lung immunity. Ten mice/group to monitor animal health. The statistical justification for group size was calculated using the SAS program to calculate the animal numbers. The analysis was carried out using a standard error of 0.5 for immunological assays, and a power of 0.9. All data are expressed as means ± SEM. Statistical significance was evaluated using Prism 9.0 software. Comparisons between two groups were analyzed by performing an unpaired Student’s t test. Comparisons between more than two groups were analyzed by performing a one-way analysis of variance (ANOVA) with Tukey’s multiple comparisons test.